---
title: "Movement tasks from a back-mounted IMU: alternatives to the XGBoost + SOM pipeline"
subtitle: "Test harness for the Zenodo dataset 10.5281/zenodo.22647205"
author: "Andrii Patrikei"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
    theme: flatly
params:
  data_dir: "data"            # where physical_exercise.csv lives / is downloaded to
  download: true              # download from Zenodo if the CSV is missing
  synthetic: false            # TRUE = fake recordings in the same CSV layout (pipeline smoke test)
  quick: true                 # TRUE = subsample children, fewer epochs / kernels / rounds
  n_children_quick: 24
  k_folds: 5                  # 9 for the full protocol (9 folds x 9 children)
  repeats: 1                  # 5 for the full protocol
  fs: 100                     # sampling rate of the sensor
  fs_low: 20                  # 4 s windows for DTW-SOM, MiniRocket, Mantis
  fs_seg: 10                  # whole segments for the variable-length DTW-SOM and the maps
  win_sec: 4                  # window length in seconds
  overlap: 0.5                # window overlap
  run_catch22: true           # needs Rcatch22
  run_dtwsom: true            # DTW-SOM on 4 s windows (needs dtwsom)
  run_dtwsom_segments: true   # DTW-SOM on whole variable-length segments (needs dtwsom)
  run_minirocket: true        # needs reticulate + python: aeon, scikit-learn
  run_mantis: false           # needs reticulate + python: mantis-tsfm, torch (downloads weights from Hugging Face)
  run_twin_test: true         # 8-label model: are the two crawls confused?
  run_maps: true              # section 7: SOM maps on all children
  plot_channel: "acc_y"       # channel drawn in the DTW-SOM maps
  som_xdim: 5
  som_ydim: 5
  use_cache: true             # reuse the feature matrix between knits (real data only)
  python: ""                  # path to the python with aeon / mantis; "" = reticulate's default
  save_figures: true          # real-data knits also write the PNGs and a results table to figures/ (used by the README)
  seed: 42
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, fig.width = 9, fig.height = 5)
set.seed(params$seed)
options(timeout = 3600)   # the CSV is 191 MB
# real-data knits keep a copy of every figure in figures/<chunk-label>-1.png for the README;
# synthetic runs don't, so fake plots never end up in the repository
if (params$save_figures && !params$synthetic) knitr::opts_chunk$set(fig.path = "figures/")
```

# What this document does

The Acta Gymnica article classified eight movement segments (seven tasks, crawling done twice)
from 81 children using hand-crafted features, XGBoost gain for feature selection and a 5x5
Self-Organizing Map. This document runs that idea again under a fixed, child-grouped evaluation
protocol and puts alternatives next to it:

| Route | Method | Package |
|---|---|---|
| Baseline | segment duration only | nnet |
| C1 | RQA (chaosrcpp) + catch22 + basic stats, XGBoost | chaosrcpp, Rcatch22, xgboost |
| C1 | same features, kohonen SOM, units and 7 super-clusters mapped to tasks | kohonen |
| C2 | DTW-SOM on raw 4 s windows | dtwsom |
| C2 | DTW-SOM on whole segments of different lengths | dtwsom |
| A | MiniRocket + ridge classifier | aeon (Python) |
| B | Mantis foundation model, frozen embeddings + random forest | mantis-tsfm (Python) |

Every model is trained on the training children only and evaluated on held-out children.
Window-level scores are averaged to one prediction per segment. Section 7 then draws the maps:
DTW-SOM prototypes of different lengths with the segments mapped to them, and the feature SOM.

Set `quick: false`, `k_folds: 9`, `repeats: 5` for the full protocol.
`synthetic: true` generates fake recordings in the same long CSV layout, so the whole pipeline,
including the loading and cleaning code, can be tested without the download. Use it first.
A real-data knit also writes every figure and the results table to `figures/`, which the README shows.

# 0. Packages and helpers

```r
install.packages(c("data.table", "xgboost", "kohonen", "nnet", "ggplot2", "MASS",
                   "Rcatch22", "reticulate", "remotes"))
remotes::install_github("andrii-patrikei/chaos-rcpp")
remotes::install_github("andrii-patrikei/dtwsom")
# Python side (the environment reticulate uses, or set params$python):
#   pip install aeon scikit-learn            # MiniRocket
#   pip install mantis-tsfm torch            # Mantis (optional)
```

```{r packages}
has <- function(p) requireNamespace(p, quietly = TRUE)
need <- c("data.table", "xgboost", "kohonen", "nnet", "ggplot2")
missing <- need[!vapply(need, has, logical(1))]
if (length(missing)) stop("Install first: ", paste(missing, collapse = ", "))

library(data.table)
library(ggplot2)

have_chaos  <- has("chaosrcpp")
have_dtwsom <- has("dtwsom")
run_catch22 <- params$run_catch22 && has("Rcatch22")
run_dtwsom     <- params$run_dtwsom && have_dtwsom
run_dtwsom_seg <- params$run_dtwsom_segments && have_dtwsom
run_maps       <- params$run_maps && have_dtwsom
if (!have_chaos) message("chaosrcpp not installed: RQA features skipped. remotes::install_github('andrii-patrikei/chaos-rcpp')")
if (params$run_catch22 && !run_catch22) message("Rcatch22 not installed: catch22 features skipped. install.packages('Rcatch22')")
if (!have_dtwsom) message("dtwsom not installed: DTW-SOM routes and maps skipped. remotes::install_github('andrii-patrikei/dtwsom')")

n_cores <- max(1L, parallel::detectCores() - 1L)
cat("cores used:", n_cores, "\n")
```

```{r python-setup}
run_py <- (params$run_minirocket || params$run_mantis) && has("reticulate")
have_aeon <- have_mantis <- FALSE
if (run_py) {
  library(reticulate)
  if (nzchar(params$python)) use_python(params$python, required = TRUE)
  run_py <- tryCatch(py_available(initialize = TRUE), error = function(e) FALSE)
  if (run_py) {
    cat("python:", py_config()$python, "\n")
    have_aeon   <- py_module_available("aeon") && py_module_available("sklearn")
    have_mantis <- py_module_available("mantis") && py_module_available("torch")
  }
}
run_minirocket <- params$run_minirocket && run_py && have_aeon
run_mantis     <- params$run_mantis && run_py && have_mantis

# what will run (anything unexpectedly "off": see the install block above, or set params$python)
status <- c("RQA features (chaosrcpp)" = have_chaos, "catch22 features" = run_catch22,
            "DTW-SOM, windows" = run_dtwsom, "DTW-SOM, segments" = run_dtwsom_seg,
            "MiniRocket (aeon)" = run_minirocket, "Mantis" = run_mantis, "SOM maps (section 7)" = run_maps)
wanted <- c(TRUE, params$run_catch22, params$run_dtwsom, params$run_dtwsom_segments,
            params$run_minirocket, params$run_mantis, params$run_maps)
cat(sprintf("%-26s %s\n", names(status), ifelse(status, "on", ifelse(wanted, "OFF (not installed)", "off"))), sep = "")
```

```{r helpers}
# ---- evaluation ------------------------------------------------------------
make_folds <- function(children, k, repeats, seed) {
  set.seed(seed)
  out <- list()
  for (r in seq_len(repeats)) {
    sh <- sample(children)
    fold_of <- rep(seq_len(k), length.out = length(sh))
    for (f in seq_len(k)) out[[length(out) + 1L]] <- list(rep = r, fold = f, test = sh[fold_of == f])
  }
  out
}

metrics <- function(truth, pred, classes) {
  truth <- factor(truth, classes); pred <- factor(pred, classes)
  cm <- table(truth, pred)
  present <- rowSums(cm) > 0
  recall <- diag(cm) / pmax(rowSums(cm), 1)
  precision <- diag(cm) / pmax(colSums(cm), 1)
  f1 <- ifelse(precision + recall > 0, 2 * precision * recall / (precision + recall), 0)
  c(accuracy = mean(truth == pred),
    balanced_accuracy = mean(recall[present]),
    macro_f1 = mean(f1[present]))
}

ari <- function(a, b) {                       # adjusted Rand index, no extra package
  tab <- table(a, b); comb2 <- function(x) x * (x - 1) / 2
  sum_ij <- sum(comb2(tab)); sum_a <- sum(comb2(rowSums(tab))); sum_b <- sum(comb2(colSums(tab)))
  expected <- sum_a * sum_b / comb2(sum(tab)); max_index <- (sum_a + sum_b) / 2
  if (max_index == expected) return(1)
  (sum_ij - expected) / (max_index - expected)
}

# average window scores per segment, return one predicted class per segment
aggregate_scores <- function(scores, seg_ids, classes) {
  scores <- as.matrix(scores)
  miss <- setdiff(classes, colnames(scores))  # a class absent from a training fold
  if (length(miss)) scores <- cbind(scores, matrix(min(scores, na.rm = TRUE) - 1, nrow(scores), length(miss),
                                                   dimnames = list(NULL, miss)))
  agg <- rowsum(scores[, classes, drop = FALSE], seg_ids)
  data.frame(seg_id = rownames(agg), pred = classes[max.col(agg, ties.method = "first")],
             stringsAsFactors = FALSE)
}

# ---- signal ----------------------------------------------------------------
decimate_block <- function(X, factor) {       # block mean (boxcar anti-alias) + subsample
  factor <- as.integer(round(factor))
  if (factor <= 1L) return(X)
  n <- (nrow(X) %/% factor) * factor
  if (n == 0L) return(matrix(colMeans(X), nrow = 1, dimnames = list(NULL, colnames(X))))
  out <- rowsum(X[seq_len(n), , drop = FALSE], rep(seq_len(n / factor), each = factor), reorder = FALSE) / factor
  rownames(out) <- NULL
  out
}

make_windows <- function(X, win, step) {
  n <- nrow(X)
  if (n < win) return(list(X[c(seq_len(n), rep(n, win - n)), , drop = FALSE]))  # pad a short segment
  starts <- seq(1L, n - win + 1L, by = step)
  lapply(starts, function(s) X[s:(s + win - 1L), , drop = FALSE])
}

# per-channel scaling fitted on training data, applied to a list of T x C matrices or an N x C x T array
fit_scaler <- function(mats) {
  all <- do.call(rbind, mats)
  list(mu = colMeans(all, na.rm = TRUE), sd = pmax(apply(all, 2, sd, na.rm = TRUE), 1e-8))
}
apply_scaler_list <- function(mats, sc) lapply(mats, function(m) sweep(sweep(m, 2, sc$mu), 2, sc$sd, "/"))
apply_scaler_array <- function(arr, sc) {
  for (ch in seq_len(dim(arr)[2])) arr[, ch, ] <- (arr[, ch, ] - sc$mu[ch]) / sc$sd[ch]
  arr
}
list_to_array <- function(mats) {             # list of T x C -> N x C x T
  arr <- array(0, dim = c(length(mats), ncol(mats[[1]]), nrow(mats[[1]])))
  for (i in seq_along(mats)) arr[i, , ] <- t(mats[[i]])
  arr
}

# ---- parallel lapply: fork on Linux / macOS, socket cluster on Windows -------
par_lapply <- function(X, FUN, export = character(0)) {
  if (n_cores == 1L || length(X) < 2L) return(lapply(X, FUN))
  if (.Platform$OS.type != "windows" && !isTRUE(getOption("movement.psock")))
    return(parallel::mclapply(X, FUN, mc.cores = n_cores))
  cl <- parallel::makePSOCKcluster(min(n_cores, length(X)))
  on.exit(parallel::stopCluster(cl), add = TRUE)
  to_global <- function(o) { if (is.function(o)) environment(o) <- globalenv(); o }
  objs <- lapply(mget(export, envir = parent.frame(), inherits = TRUE), to_global)
  parallel::clusterCall(cl, function(o) { list2env(o, envir = globalenv()); NULL }, objs)
  parallel::parLapply(cl, X, to_global(FUN))
}

class_palette <- function(classes) setNames(hcl.colors(length(classes), "Dark 3"), classes)
```

# 1. Data

## 1.1 Load

```{r user-settings}
# If the automatic column detection below gets it wrong, fill these in and re-knit.
col_override <- list(id = NULL, task = NULL, time = NULL,
                     acc = NULL,     # e.g. c("acc_x", "acc_y", "acc_z")
                     gyro = NULL)    # e.g. c("gyro_x", "gyro_y", "gyro_z")
crawl_override <- NULL   # the crawling label; if the two crawls have different labels give both, c("2", "8")
task_names <- NULL       # optional readable names for the plots, e.g. c("1" = "run", "2" = "crawl", ...)
```

```{r synthetic-generator}
# Fake recordings in a long CSV layout: 8 stations per child (crawling at stations 2 and 8),
# unlabeled walks between stations, random sensor dropouts, one segment that is 40% missing.
make_synthetic_raw <- function(n_children = 24, fs = 100, seed = 1) {
  set.seed(seed)
  freq <- c(2.8, 0.8, 2.2, 1.6, 0.5, 0.4, 3.2, 0.8)
  amp  <- c(3.0, 0.6, 2.5, 2.0, 1.2, 1.0, 4.0, 0.6)
  base_dur <- c(12, 20, 15, 10, 25, 18, 8, 20)
  parts <- list()
  for (c in seq_len(n_children)) {
    t0 <- 0; id <- sprintf("P%03d", c)
    for (k in 1:8) {
      dur <- base_dur[k] * runif(1, 0.7, 1.4); n <- round(dur * fs); t <- seq_len(n) / fs
      f <- freq[k] * runif(1, 0.85, 1.15); a <- amp[k] * runif(1, 0.8, 1.2) * (1 + 0.15 * (k == 8))
      X <- cbind(a * sin(2 * pi * f * t) + rnorm(n, 0, 0.3),
                 1 + 0.5 * a * cos(2 * pi * f * t) + rnorm(n, 0, 0.3),
                 0.3 * a * sin(4 * pi * f * t + 1) + rnorm(n, 0, 0.3),
                 50 * a * cos(2 * pi * f * t) + rnorm(n, 0, 10),
                 20 * a * sin(2 * pi * f * t + 0.5) + rnorm(n, 0, 10),
                 10 * a * sin(6 * pi * f * t) + rnorm(n, 0, 10))
      if (runif(1) < 0.3) {
        X[cbind(sample(n, 10), sample(6, 10, replace = TRUE))] <- NA
        s0 <- sample(n - 40, 1); X[s0:(s0 + 20), ] <- NA
      }
      parts[[length(parts) + 1L]] <- data.table(participant = id, station = k, timestamp = t0 + t,
                                                 ax = X[, 1], ay = X[, 2], az = X[, 3],
                                                 gx = X[, 4], gy = X[, 5], gz = X[, 6])
      t0 <- t0 + dur
      if (k < 8) {
        m <- round(0.5 * fs)
        parts[[length(parts) + 1L]] <- data.table(participant = id, station = NA_integer_, timestamp = t0 + seq_len(m) / fs,
                                                   ax = rnorm(m), ay = 1 + rnorm(m), az = rnorm(m),
                                                   gx = rnorm(m, 0, 30), gy = rnorm(m, 0, 30), gz = rnorm(m, 0, 30))
        t0 <- t0 + m / fs
      }
    }
  }
  raw <- rbindlist(parts)
  bad <- raw[participant == "P001" & station == 5L, which = TRUE]
  raw[bad[seq_len(round(0.4 * length(bad)))], c("ax", "ay", "az", "gx", "gy", "gz") := NA_real_]
  raw
}
```

```{r download, eval = !params$synthetic}
zenodo_url <- "https://zenodo.org/records/22647205/files/physical_exercise.csv?download=1"
zenodo_md5 <- "5ea4d80cbcfdaa1e54491821db57ab74"
csv_path <- file.path(params$data_dir, "physical_exercise.csv")
dir.create(params$data_dir, showWarnings = FALSE, recursive = TRUE)
if (!file.exists(csv_path)) {
  if (!params$download) stop("CSV not found at ", csv_path, " and download = FALSE")
  message("Downloading 191 MB from Zenodo ...")
  download.file(zenodo_url, csv_path, mode = "wb")
}
cat("md5 matches the Zenodo record:", identical(unname(tools::md5sum(csv_path)), zenodo_md5), "\n")
```

```{r load}
if (params$synthetic) {
  raw <- make_synthetic_raw(n_children = params$n_children_quick, fs = params$fs, seed = params$seed)
  if (is.null(task_names)) task_names <- c("1" = "run", "2" = "crawl", "3" = "shuttle", "4" = "rings",
                                           "5" = "balance", "6" = "hoops", "7" = "bench")
} else {
  raw <- fread(csv_path)
}
str(raw)
print(head(raw))

guess_columns <- function(dt) {
  nm <- names(dt); low <- tolower(nm)
  nun <- vapply(dt, function(v) length(unique(v)), integer(1))
  pick <- function(pattern, exclude = character(0)) {
    hit <- setdiff(nm[grepl(pattern, low)], exclude); if (length(hit)) hit else NULL
  }
  time <- pick("time|stamp|^t$")
  acc  <- pick("^acc|^a[xyz]$|accel", time)
  gyro <- pick("^gyr|^g[xyz]$|gyro|rot", time)
  id   <- pick("child|subject|participant|proband|person|^id$", c(time, acc, gyro))
  task <- pick("task|activity|exercise|segment|station|label|class", c(time, acc, gyro, id))
  if (is.null(id))   id   <- nm[which(nun > 50 & nun < 200 & !(nm %in% c(acc, gyro, time)))][1]
  if (is.null(task)) task <- nm[which(nun >= 6 & nun <= 12 & !(nm %in% c(id, acc, gyro, time)))][1]
  list(id = id[1], task = task[1], time = if (length(time)) time[1] else NULL,
       acc = head(acc, 3), gyro = head(gyro, 3))
}
cols <- guess_columns(raw)
for (k in names(col_override)) if (!is.null(col_override[[k]])) cols[[k]] <- col_override[[k]]
str(cols)
stopifnot(!is.null(cols$id), !is.null(cols$task), length(cols$acc) == 3, length(cols$gyro) == 3)
```

## 1.2 Segments

Rows are kept in file order within each child. Each child's recording is cut into consecutive
runs of the task label (`rleid`), which gives the 8 segments. The two crawls share one class,
so the classification target has 7 classes; `seg_label` keeps them apart for the twin test.
Short gaps inside a segment are filled by linear interpolation; segments that are mostly
missing are dropped and listed.

```{r segments}
dt <- raw[, c(cols$id, cols$task, cols$time, cols$acc, cols$gyro), with = FALSE]
setnames(dt, c("child", "task", if (!is.null(cols$time)) "time", "acc_x", "acc_y", "acc_z", "gyr_x", "gyr_y", "gyr_z"))
rm(raw); invisible(gc())
sensor_cols <- c("acc_x", "acc_y", "acc_z", "gyr_x", "gyr_y", "gyr_z")
for (v in sensor_cols) if (!is.numeric(dt[[v]])) set(dt, j = v, value = suppressWarnings(as.numeric(dt[[v]])))
dt[, child := trimws(as.character(child))]
dt[, task := trimws(as.character(task))]
setorder(dt, child)                            # stable: keeps the file order inside each child

if ("time" %in% names(dt) && (is.numeric(dt$time) || inherits(dt$time, "POSIXt"))) {
  d <- dt[, .(d = if (inherits(time, "POSIXt")) as.numeric(diff(time), units = "secs") else as.numeric(diff(time))), by = child]$d
  d <- d[is.finite(d)]
  cat(sprintf("median time step: %.5g (file units; %.4f s expected at %d Hz); backward steps: %d\n",
              median(d[d > 0]), 1 / params$fs, params$fs, sum(d < 0)))
}

cat("rows:", nrow(dt), "\nmissing / non-finite values per sensor column:\n")
print(dt[, lapply(.SD, function(v) sum(!is.finite(v))), .SDcols = sensor_cols])
cat("rows without child id:", dt[, sum(is.na(child) | child == "")],
    "  rows without task label:", dt[, sum(is.na(task) | task == "" | task == "NA")], "\n")
dt <- dt[!is.na(child) & child != "" & !is.na(task) & task != "" & task != "NA"]
dt[, seg := rleid(task), by = child]

# which label is crawling?
task_levels <- sort(unique(dt$task))
cat("task labels in the file:", paste(task_levels, collapse = ", "), "\n")
crawl_class <- NULL
if (!is.null(crawl_override)) {
  crawl_class <- crawl_override[1]
  if (length(crawl_override) == 2) dt[task == crawl_override[2], task := crawl_override[1]]
} else {
  hit <- task_levels[grepl("crawl|plaz|pliz|pl\u00ed\u017e|lez", tolower(task_levels))]
  if (length(hit) == 1) {
    crawl_class <- hit
  } else if (length(hit) == 2) {
    crawl_class <- hit[1]; dt[task == hit[2], task := hit[1]]
  } else if (all(grepl("^[0-9]+$", task_levels)) && all(c("2", "8") %in% task_levels)) {
    crawl_class <- "2"; dt[task == "8", task := "2"]          # numeric station codes: 8 is the second crawl
  }
}
if (!is.null(task_names)) {
  dt[task %in% names(task_names), task := unname(task_names[task])]
  if (!is.null(crawl_class) && crawl_class %in% names(task_names)) crawl_class <- unname(task_names[crawl_class])
}
cat("crawling label:", if (is.null(crawl_class)) "NOT DETECTED, set crawl_override" else crawl_class, "\n")

seg_tab <- dt[, .(task = task[1], n = .N,
                  acc_x = list(acc_x), acc_y = list(acc_y), acc_z = list(acc_z),
                  gyr_x = list(gyr_x), gyr_y = list(gyr_y), gyr_z = list(gyr_z)),
              by = .(child, seg)]
rm(dt); invisible(gc())
seg_tab[, dur_s := n / params$fs]
seg_tab[, class := task]
seg_tab[, seg_label := class]
if (!is.null(crawl_class)) seg_tab[class == crawl_class, seg_label := paste0(class, "_", seq_len(.N)), by = child]
seg_tab[, seg_id := paste(child, seg, sep = "_")]

fill_gaps <- function(X) {                    # linear interpolation per channel, ends held constant
  for (j in seq_len(ncol(X))) {
    v <- X[, j]; ok <- is.finite(v)
    if (!any(ok)) return(NULL)
    if (!all(ok)) X[, j] <- approx(which(ok), v[ok], xout = seq_along(v), rule = 2)$y
  }
  X
}
max_na_frac <- 0.2                            # drop a segment if more than 20% of its values are missing
min_len <- params$fs                          # or if it is shorter than 1 s

segments <- lapply(seq_len(nrow(seg_tab)), function(i) {
  X <- cbind(acc_x = seg_tab$acc_x[[i]], acc_y = seg_tab$acc_y[[i]], acc_z = seg_tab$acc_z[[i]],
             gyr_x = seg_tab$gyr_x[[i]], gyr_y = seg_tab$gyr_y[[i]], gyr_z = seg_tab$gyr_z[[i]])
  na_frac <- mean(!is.finite(X))
  Xf <- if (na_frac <= max_na_frac && nrow(X) >= min_len) fill_gaps(X) else NULL
  list(seg_id = seg_tab$seg_id[i], child = seg_tab$child[i], class = seg_tab$class[i],
       seg_label = seg_tab$seg_label[i], X = Xf, na_frac = na_frac)
})
seg_tab[, na_frac := vapply(segments, `[[`, numeric(1), "na_frac")]
kept <- !vapply(segments, function(s) is.null(s$X), logical(1))
cat(sprintf("segments: %d; with gaps: %d; interpolated: %d; dropped (>%.0f%% missing or < 1 s): %d\n",
            nrow(seg_tab), sum(seg_tab$na_frac > 0), sum(seg_tab$na_frac > 0 & kept), 100 * max_na_frac, sum(!kept)))
if (any(!kept)) print(seg_tab[!kept, .(child, seg, task, n, na_frac = round(na_frac, 3))])

# drop labels that are far rarer than the others (stray or mistyped labels)
cnt <- seg_tab[kept, .N, by = class]
rare <- cnt[N < 0.2 * median(N), class]
if (length(rare)) { cat("dropping rare labels:", paste(rare, collapse = ", "), "\n"); kept <- kept & !(seg_tab$class %in% rare) }

segments <- segments[kept]
seg_meta <- seg_tab[kept, .(seg_id, child, seg, class, seg_label, n, dur_s, na_frac)]
rm(seg_tab); invisible(gc())
stopifnot(all(vapply(segments, function(s) all(is.finite(s$X)), logical(1))),
          identical(vapply(segments, `[[`, "", "seg_id"), seg_meta$seg_id))
```

## 1.3 Audit

```{r audit}
classes <- sort(unique(seg_meta$class))
pal <- class_palette(classes)
cat("children:", length(unique(seg_meta$child)), " segments:", nrow(seg_meta), " classes:", length(classes), "\n")
print(table(segments_per_child = seg_meta[, .N, by = child]$N))   # expect 8
print(seg_meta[, .(n = .N, median_dur_s = median(dur_s), min_dur_s = min(dur_s), max_dur_s = max(dur_s),
                   with_gaps = sum(na_frac > 0)), by = class][order(class)])

ggplot(seg_meta, aes(x = class, y = dur_s, fill = class)) + geom_boxplot(show.legend = FALSE) +
  scale_fill_manual(values = pal) + labs(title = "Segment duration by task", y = "seconds", x = NULL) + theme_minimal()

rng <- do.call(rbind, lapply(segments[seq_len(min(50, length(segments)))], function(s) apply(s$X, 2, range)))
cat("value ranges over the first 50 segments (check units: g vs m/s^2, deg/s):\n")
print(round(apply(rng, 2, range), 2))
```

```{r subsample}
if (params$quick && !params$synthetic) {
  keep <- sort(sample(unique(seg_meta$child), min(params$n_children_quick, length(unique(seg_meta$child)))))
  sel <- seg_meta$child %in% keep
  segments <- segments[sel]; seg_meta <- seg_meta[sel]
  cat("quick mode: using", length(keep), "children,", nrow(seg_meta), "segments\n")
}
children <- sort(unique(seg_meta$child))
folds <- make_folds(children, k = params$k_folds, repeats = params$repeats, seed = params$seed)
cat(length(folds), "folds\n")
```

# 2. Windows and whole segments

Windows of `r params$win_sec` s with `r params$overlap * 100`% overlap. Features use the
100 Hz windows; DTW-SOM, MiniRocket and Mantis use the same windows at `r params$fs_low` Hz.
The whole segments are also kept at `r params$fs_seg` Hz, each with its own length, for the
variable-length DTW-SOM.

```{r windows}
win_hi <- params$win_sec * params$fs
step_hi <- max(1L, round(win_hi * (1 - params$overlap)))
dec_lo <- params$fs / params$fs_low
dec_seg <- params$fs / params$fs_seg
stopifnot(dec_lo == round(dec_lo), dec_seg == round(dec_seg))

ws_all <- lapply(segments, function(s) make_windows(s$X, win_hi, step_hi))
n_w <- lengths(ws_all)
win_list <- unlist(ws_all, recursive = FALSE); rm(ws_all)
win_meta <- data.table(seg_id = rep(seg_meta$seg_id, n_w), child = rep(seg_meta$child, n_w),
                       class = rep(seg_meta$class, n_w), seg_label = rep(seg_meta$seg_label, n_w),
                       w = sequence(n_w))
win_meta[, win_id := .I]
win_lo <- lapply(win_list, decimate_block, factor = dec_lo)
seg_lo <- lapply(segments, function(s) decimate_block(s$X, dec_seg))
names(seg_lo) <- seg_meta$seg_id
if (run_minirocket || run_mantis) arr_lo <- list_to_array(win_lo)
cat(sprintf("%d windows of %d samples (%d at %d Hz); %d whole segments of %d to %d samples at %d Hz\n",
            nrow(win_meta), win_hi, nrow(win_lo[[1]]), params$fs_low, length(seg_lo),
            min(vapply(seg_lo, nrow, 1L)), max(vapply(seg_lo, nrow, 1L)), params$fs_seg))
```

# 3. Features (route C1)

RQA follows the article's design: the three accelerometer axes form the state vector (no delay
embedding), likewise the three gyroscope axes. What changes: the threshold is chosen per window
for a fixed recurrence rate (5%) instead of a fixed distance, and a Theiler window excludes the
main diagonal. catch22 gives 22 general-purpose features per channel.

```{r features}
basic_block <- function(v, prefix) {
  d <- diff(v)
  out <- c(mean = mean(v), sd = sd(v), min = min(v), max = max(v),
           rms = sqrt(mean(v^2)), mad_diff = mean(abs(d)), zcr = mean(diff(sign(v - mean(v))) != 0))
  setNames(out, paste0(prefix, "_", names(out)))
}
rqa_names <- if (have_chaos) names(unlist(chaosrcpp::rqa_features(matrix(rnorm(300), 100, 3), rr = 0.05))) else character(0)
rqa_block <- function(M, prefix) {
  f <- tryCatch(unlist(chaosrcpp::rqa_features(M, rr = 0.05, theiler = 1L, lmin = 2L, vmin = 2L)),
                error = function(e) NULL)
  if (length(f) != length(rqa_names)) f <- rep(NA_real_, length(rqa_names))
  setNames(as.numeric(f), paste0(prefix, "_", rqa_names))
}
catch22_names <- if (run_catch22) suppressWarnings(Rcatch22::catch22_all(rnorm(200)))[["names"]] else character(0)
catch22_block <- function(v, prefix) {
  r <- tryCatch(suppressWarnings(Rcatch22::catch22_all(v)), error = function(e) NULL)
  vals <- if (is.null(r)) rep(NA_real_, length(catch22_names)) else as.numeric(r[["values"]])[match(catch22_names, r[["names"]])]
  setNames(vals, paste0(prefix, "_", catch22_names))
}
featurize_window <- function(W) {
  acc <- W[, 1:3]; gyr <- W[, 4:6]
  ch <- cbind(W, acc_mag = sqrt(rowSums(acc^2)), gyr_mag = sqrt(rowSums(gyr^2)))
  out <- unlist(lapply(colnames(ch), function(nm) basic_block(ch[, nm], nm)))
  if (have_chaos) out <- c(out, rqa_block(acc, "rqa_acc"), rqa_block(gyr, "rqa_gyr"))
  if (run_catch22) out <- c(out, unlist(lapply(colnames(ch), function(nm) catch22_block(ch[, nm], nm))))
  out
}

feat_file <- file.path(params$data_dir, sprintf("cache_features_%s_seed%d_%dchildren_%dwin_%gs_%g_rqa%d_c22%d.rds",
                       if (params$synthetic) "synthetic" else "real", params$seed, length(children), nrow(win_meta),
                       params$win_sec, params$overlap, have_chaos, run_catch22))
if (params$use_cache && !params$synthetic && file.exists(feat_file)) {
  feat <- readRDS(feat_file); cat("features loaded from", feat_file, "\n")
} else {
  t0 <- Sys.time()
  feat_list <- par_lapply(win_list, featurize_window,
                          export = c("featurize_window", "basic_block", "rqa_block", "catch22_block",
                                     "rqa_names", "catch22_names", "have_chaos", "run_catch22"))
  stopifnot(length(unique(lengths(feat_list))) == 1)
  feat <- do.call(rbind, feat_list); rm(feat_list)
  feat[!is.finite(feat)] <- NA
  cat(sprintf("%d windows x %d features in %.1f s\n", nrow(feat), ncol(feat), as.numeric(Sys.time() - t0, units = "secs")))
  if (params$use_cache && !params$synthetic) { dir.create(params$data_dir, showWarnings = FALSE); saveRDS(feat, feat_file) }
}
stopifnot(nrow(feat) == nrow(win_meta))
cat("features with any missing value:", sum(colSums(is.na(feat)) > 0), "of", ncol(feat), "\n")

# segment-level features (mean over the windows of a segment) for the feature map in section 7
ok <- !is.na(feat)
feat_seg <- rowsum(ifelse(ok, feat, 0), win_meta$seg_id) / pmax(rowsum(ok * 1, win_meta$seg_id), 1)
feat_seg <- feat_seg[seg_meta$seg_id, , drop = FALSE]
```

# 4. Models

Every model function takes training and test data and returns a score matrix with one column
per class, which is then averaged per segment.

```{r model-functions}
fit_duration <- function(seg_tr, seg_te, classes) {
  y <- droplevels(factor(seg_tr$class, levels = classes))
  m <- nnet::multinom(y ~ log(dur_s), data = data.frame(y = y, dur_s = seg_tr$dur_s), trace = FALSE, maxit = 500)
  p <- predict(m, newdata = data.frame(dur_s = seg_te$dur_s), type = "probs")
  if (!is.matrix(p)) p <- matrix(p, nrow = nrow(seg_te), byrow = TRUE, dimnames = list(NULL, levels(y)))
  out <- matrix(0, nrow(seg_te), length(classes), dimnames = list(NULL, classes))
  out[, colnames(p)] <- p
  out
}

xgb_params <- function(K) list(objective = "multi:softprob", num_class = K, eta = 0.05, max_depth = 4,
                               subsample = 0.8, colsample_bytree = 0.8, min_child_weight = 2, nthread = n_cores)
fit_xgb <- function(Xtr, ytr, Xte, classes, nrounds = 300, top_k = NULL) {
  y_i <- as.integer(factor(ytr, classes)) - 1L
  train <- function(X) xgboost::xgb.train(params = xgb_params(length(classes)),
                                          data = xgboost::xgb.DMatrix(X, label = y_i), nrounds = nrounds, verbose = 0)
  m <- train(Xtr)
  if (!is.null(top_k)) {                     # the article's selection step, but inside the fold
    keep <- head(xgboost::xgb.importance(model = m)$Feature, top_k)
    Xtr <- Xtr[, keep, drop = FALSE]; Xte <- Xte[, keep, drop = FALSE]
    m <- train(Xtr)
  }
  p <- predict(m, xgboost::xgb.DMatrix(Xte))
  if (!is.matrix(p)) p <- matrix(p, ncol = length(classes), byrow = TRUE)
  colnames(p) <- classes
  list(scores = p, model = m)
}

# map SOM units to classes by majority of the training data; empty units inherit the nearest unit's class
unit_to_class <- function(unit_tr, ytr, D_units, classes) {
  uc <- rep(NA_character_, nrow(D_units))
  maj <- tapply(ytr, unit_tr, function(v) names(which.max(table(factor(v, classes)))))
  uc[as.integer(names(maj))] <- maj
  for (u in which(is.na(uc))) { cand <- which(!is.na(uc)); uc[u] <- uc[cand[which.min(D_units[u, cand])]] }
  uc
}
one_hot <- function(lab, classes) {
  m <- matrix(0, length(lab), length(classes), dimnames = list(NULL, classes))
  m[cbind(seq_along(lab), match(lab, classes))] <- 1
  m
}
super_clusters <- function(D, k) cutree(hclust(as.dist(D), "ward.D2"), k = min(k, nrow(D)))

fit_som <- function(Ftr, ytr, Fte, yte, classes, xdim = params$som_xdim, ydim = params$som_ydim, rlen = 200) {
  mu <- colMeans(Ftr, na.rm = TRUE); sdv <- apply(Ftr, 2, sd, na.rm = TRUE); sdv[!is.finite(sdv) | sdv == 0] <- 1
  Ztr <- scale(Ftr, mu, sdv); Zte <- scale(Fte, mu, sdv); Ztr[!is.finite(Ztr)] <- 0; Zte[!is.finite(Zte)] <- 0
  s <- kohonen::som(Ztr, grid = kohonen::somgrid(xdim, ydim, "hexagonal"), rlen = rlen, mode = "batch")
  codes <- s$codes[[1]]; D <- as.matrix(dist(codes))
  unit_tr <- s$unit.classif; unit_te <- kohonen::map(s, Zte)$unit.classif
  uc <- unit_to_class(unit_tr, ytr, D, classes)
  super <- super_clusters(D, length(classes))
  cent <- t(sapply(sort(unique(super)), function(k) colMeans(codes[super == k, , drop = FALSE])))
  sc <- unit_to_class(super[unit_tr], ytr, as.matrix(dist(cent)), classes)
  list(scores_unit = one_hot(uc[unit_te], classes), scores_super = one_hot(sc[super[unit_te]], classes),
       ari_unit = ari(yte, unit_te), ari_super = ari(yte, super[unit_te]), som = s)
}

fit_dtwsom <- function(Ltr, ytr, Lte, yte, classes, xdim = params$som_xdim, ydim = params$som_ydim,
                       epochs = 15, window = NULL, seed = 1) {
  while (xdim * ydim > length(Ltr) && xdim > 2) { xdim <- xdim - 1L; ydim <- ydim - 1L }
  if (is.null(window)) window <- max(2L, round(0.1 * median(vapply(Ltr, nrow, 1L))))
  fit <- dtwsom::dtw_som(Ltr, xdim = xdim, ydim = ydim, topo = "hexagonal", epochs = epochs, mode = "batch",
                         window = window, smooth = 3, normalize = "none", threads = n_cores,
                         seed = seed, verbose = FALSE)
  D <- dtwsom::dtw_pairwise(fit$codes, window = window, threads = n_cores)
  unit_tr <- fit$unit.classif
  unit_te <- predict(fit, Lte, threads = n_cores)$bmu
  uc <- unit_to_class(unit_tr, ytr, D, classes)
  super <- super_clusters(D, length(classes)); ks <- max(super)
  Dsup <- matrix(0, ks, ks)
  for (a in seq_len(ks)) for (b in seq_len(ks)) Dsup[a, b] <- mean(D[super == a, super == b, drop = FALSE])
  sc <- unit_to_class(super[unit_tr], ytr, Dsup, classes)
  list(scores_unit = one_hot(uc[unit_te], classes), scores_super = one_hot(sc[super[unit_te]], classes),
       ari_unit = ari(yte, unit_te), ari_super = ari(yte, super[unit_te]), fit = fit)
}
```

```{python minirocket-py, eval = run_minirocket}
import numpy as np
from aeon.transformations.collection.convolution_based import MiniRocket
from sklearn.linear_model import RidgeClassifierCV

def minirocket_fit_predict(Xtr, ytr, Xte, n_kernels=10000, seed=0, n_jobs=1):
    Xtr = np.ascontiguousarray(Xtr, dtype=np.float32)
    Xte = np.ascontiguousarray(Xte, dtype=np.float32)
    tf = MiniRocket(n_kernels=int(n_kernels), random_state=int(seed), n_jobs=int(n_jobs))
    Ftr = tf.fit_transform(Xtr)
    Fte = tf.transform(Xte)
    clf = RidgeClassifierCV(alphas=np.logspace(-3, 3, 10)).fit(Ftr, np.asarray(ytr))
    return {"scores": clf.decision_function(Fte), "classes": [str(c) for c in clf.classes_]}
```

```{python mantis-py, eval = run_mantis}
import numpy as np
import torch
import torch.nn.functional as F
from mantis.architecture import MantisV1, MantisV2
from mantis.trainer import MantisTrainer
from sklearn.ensemble import RandomForestClassifier

def mantis_embed(X, checkpoint="paris-noah/MantisV2", layer=2, device="cpu", batch_size=256, n_threads=1):
    """Frozen encoder. X: (n, channels, T). Each channel is embedded separately and concatenated."""
    torch.set_num_threads(int(n_threads))
    cls = MantisV2 if "V2" in checkpoint else MantisV1
    net = cls(device=device, return_transf_layer=int(layer), output_token="combined").from_pretrained(checkpoint)
    model = MantisTrainer(device=device, network=net)
    X = np.ascontiguousarray(X, dtype=np.float32)
    Xr = F.interpolate(torch.tensor(X), size=512, mode="linear", align_corners=False).numpy()
    return model.transform(Xr, batch_size=int(batch_size))

def rf_fit_predict(Ztr, ytr, Zte, seed=0, n_jobs=1):
    clf = RandomForestClassifier(n_estimators=300, random_state=int(seed), n_jobs=int(n_jobs))
    clf.fit(np.asarray(Ztr), np.asarray(ytr))
    return {"scores": clf.predict_proba(np.asarray(Zte)), "classes": [str(c) for c in clf.classes_]}
```

The Mantis encoder is frozen, so the embeddings are computed once. They use one global scaling
per channel computed from all windows without labels; the random forest is trained inside each fold.

```{r mantis-embed, eval = run_mantis}
t0 <- Sys.time()
Z_mantis <- py$mantis_embed(apply_scaler_array(arr_lo, fit_scaler(win_lo)), n_threads = n_cores)
cat(sprintf("Mantis embeddings: %d x %d in %.0f s\n", nrow(Z_mantis), ncol(Z_mantis), as.numeric(Sys.time() - t0, units = "secs")))
```

# 5. Cross-validation

```{r cv-loop}
quick <- params$quick
results <- list(); cluster_quality <- list(); twin_results <- list(); timing <- list()

add_result <- function(method, scores, seg_ids, f) {
  ag <- aggregate_scores(scores, seg_ids, classes)
  ag$method <- method; ag$rep <- f$rep; ag$fold <- f$fold
  results[[length(results) + 1L]] <<- ag
}
add_quality <- function(method, fit, f) {
  cluster_quality[[length(cluster_quality) + 1L]] <<- data.frame(method = method, rep = f$rep, fold = f$fold,
                                                                 ari_unit = fit$ari_unit, ari_super = fit$ari_super)
}
timed <- function(method, f, expr) {
  t0 <- Sys.time(); val <- expr
  timing[[length(timing) + 1L]] <<- data.frame(method = method, rep = f$rep, fold = f$fold,
                                               seconds = as.numeric(Sys.time() - t0, units = "secs"))
  val
}

for (fi in seq_along(folds)) {
  f <- folds[[fi]]
  te_idx <- which(win_meta$child %in% f$test); tr_idx <- which(!(win_meta$child %in% f$test))
  ytr <- win_meta$class[tr_idx]; yte <- win_meta$class[te_idx]; seg_te <- win_meta$seg_id[te_idx]
  s_tr <- which(!(seg_meta$child %in% f$test)); s_te <- which(seg_meta$child %in% f$test)
  cat(sprintf("rep %d fold %d: %d train / %d test windows, %d / %d segments, %d test children\n",
              f$rep, f$fold, length(tr_idx), length(te_idx), length(s_tr), length(s_te), length(f$test)))

  ## baseline: duration only
  add_result("Duration only", fit_duration(seg_meta[s_tr], seg_meta[s_te], classes), seg_meta$seg_id[s_te], f)

  ## C1: features -> XGBoost (all features, and the article's top-27 chosen inside the fold)
  nr <- if (quick) 150 else 400
  xg <- timed("XGBoost, all features", f, fit_xgb(feat[tr_idx, , drop = FALSE], ytr, feat[te_idx, , drop = FALSE], classes, nrounds = nr))
  add_result("XGBoost, all features", xg$scores, seg_te, f)
  xg27 <- timed("XGBoost, top-27 in-fold", f, fit_xgb(feat[tr_idx, , drop = FALSE], ytr, feat[te_idx, , drop = FALSE], classes, nrounds = nr, top_k = 27))
  add_result("XGBoost, top-27 in-fold", xg27$scores, seg_te, f)

  ## C1: features -> kohonen SOM
  so <- timed("SOM features", f, fit_som(feat[tr_idx, , drop = FALSE], ytr, feat[te_idx, , drop = FALSE], yte, classes,
                                            rlen = if (quick) 100 else 300))
  add_result("SOM features, unit majority", so$scores_unit, seg_te, f)
  add_result("SOM features, 7 super-clusters", so$scores_super, seg_te, f)
  add_quality("SOM features (windows)", so, f)

  ## C2: DTW-SOM on the raw 4 s windows
  if (run_dtwsom) {
    sc <- fit_scaler(win_lo[tr_idx])
    ds <- timed("DTW-SOM windows", f, fit_dtwsom(apply_scaler_list(win_lo[tr_idx], sc), ytr, apply_scaler_list(win_lo[te_idx], sc), yte,
                                                 classes, epochs = if (quick) 10 else 30, seed = params$seed + fi))
    add_result("DTW-SOM windows, unit majority", ds$scores_unit, seg_te, f)
    add_result("DTW-SOM windows, 7 super-clusters", ds$scores_super, seg_te, f)
    add_quality("DTW-SOM (windows)", ds, f)
  }

  ## C2: DTW-SOM on whole segments of different lengths
  if (run_dtwsom_seg) {
    sc <- fit_scaler(seg_lo[s_tr])
    ws <- max(2L, round(0.1 * median(vapply(seg_lo[s_tr], nrow, 1L))))
    dss <- timed("DTW-SOM segments", f, fit_dtwsom(apply_scaler_list(seg_lo[s_tr], sc), seg_meta$class[s_tr],
                                                   apply_scaler_list(seg_lo[s_te], sc), seg_meta$class[s_te], classes,
                                                   epochs = if (quick) 15 else 40, window = ws, seed = params$seed + fi))
    add_result("DTW-SOM segments, unit majority", dss$scores_unit, seg_meta$seg_id[s_te], f)
    add_result("DTW-SOM segments, 7 super-clusters", dss$scores_super, seg_meta$seg_id[s_te], f)
    add_quality("DTW-SOM (segments)", dss, f)
  }

  ## A: MiniRocket + ridge
  if (run_minirocket) {
    sc <- fit_scaler(win_lo[tr_idx])
    mr <- timed("MiniRocket + ridge", f, py$minirocket_fit_predict(apply_scaler_array(arr_lo[tr_idx, , , drop = FALSE], sc), ytr,
                                                                   apply_scaler_array(arr_lo[te_idx, , , drop = FALSE], sc),
                                                                   n_kernels = if (quick) 2000 else 10000, seed = params$seed, n_jobs = n_cores))
    add_result("MiniRocket + ridge", `colnames<-`(mr$scores, unlist(mr$classes)), seg_te, f)
  }

  ## B: Mantis frozen embeddings + random forest
  if (run_mantis) {
    ma <- timed("Mantis frozen + RF", f, py$rf_fit_predict(Z_mantis[tr_idx, , drop = FALSE], ytr, Z_mantis[te_idx, , drop = FALSE],
                                                           seed = params$seed, n_jobs = n_cores))
    add_result("Mantis frozen + RF", `colnames<-`(ma$scores, unlist(ma$classes)), seg_te, f)
  }

  ## twin test: can an 8-label model tell the first crawl from the second?
  if (params$run_twin_test && !is.null(crawl_class)) {
    labs8 <- sort(unique(win_meta$seg_label))
    tw <- fit_xgb(feat[tr_idx, , drop = FALSE], win_meta$seg_label[tr_idx], feat[te_idx, , drop = FALSE], labs8, nrounds = nr)
    ag <- aggregate_scores(tw$scores, seg_te, labs8); ag$rep <- f$rep; ag$fold <- f$fold
    twin_results[[length(twin_results) + 1L]] <- ag
  }
}
```

# 6. Results

```{r results-table}
res <- rbindlist(results, use.names = TRUE)
res <- merge(res, seg_meta[, .(seg_id, child, class)], by = "seg_id")
per_fold <- res[, as.list(metrics(class, pred, classes)), by = .(method, rep, fold)]
summary_tab <- per_fold[, .(balanced_accuracy = mean(balanced_accuracy), sd = sd(balanced_accuracy),
                            macro_f1 = mean(macro_f1), folds = .N), by = method]
if (length(timing)) {                        # one fit serves both mappings of a SOM: match on the name before the comma
  tim <- rbindlist(timing)[, .(fit_seconds = round(mean(seconds), 1)), by = .(key = method)]
  summary_tab[, key := ifelse(method %in% tim$key, method, sub(",.*$", "", method))]
  summary_tab <- merge(summary_tab, tim, by = "key", all.x = TRUE)[, key := NULL]
}
setorder(summary_tab, -balanced_accuracy)
if (params$save_figures && !params$synthetic) {
  dir.create("figures", showWarnings = FALSE)
  writeLines(c("# Results of the latest run", "",
               sprintf("%s: %d children, %d folds x %d repeats, quick = %s. Balanced accuracy and macro F1 on held-out children.",
                       format(Sys.Date()), length(children), params$k_folds, params$repeats, params$quick), "",
               knitr::kable(summary_tab, format = "pipe", digits = 3)), "figures/results_table.md")
}
knitr::kable(summary_tab, digits = 3, caption = "Segment-level performance on held-out children (mean over folds)")
```

```{r results-plot, fig.height = 6}
per_fold[, method := factor(method, levels = rev(summary_tab$method))]
ggplot(per_fold, aes(x = method, y = balanced_accuracy)) +
  geom_boxplot(outlier.shape = NA, fill = "grey92") + geom_jitter(width = 0.15, height = 0, alpha = 0.6) +
  coord_flip() + labs(title = "Balanced accuracy per fold (held-out children)", x = NULL, y = "balanced accuracy") +
  theme_minimal()
```

## Clustering quality

Adjusted Rand index between the tasks of the held-out data and (a) the SOM unit they land on,
(b) the super-cluster. This is the comparison closest to the article's own question.

```{r cluster-quality}
if (length(cluster_quality)) {
  cq <- rbindlist(cluster_quality)[, .(ari_unit = mean(ari_unit), ari_super = mean(ari_super), folds = .N), by = method]
  knitr::kable(cq, digits = 3)
}
```

## Confusion matrix of the best method

```{r confusion, fig.height = 5.5}
best <- as.character(summary_tab$method[1])
cm <- res[method == best, table(truth = factor(class, levels = classes), pred = factor(pred, levels = classes))]
cat("best:", best, "\n"); print(cm)
ggplot(as.data.frame(cm), aes(x = pred, y = truth, fill = Freq)) + geom_tile() + geom_text(aes(label = Freq)) +
  scale_fill_gradient(low = "white", high = "steelblue") +
  labs(title = paste("Confusion,", best), x = "predicted", y = "true") +
  theme_minimal() + theme(axis.text.x = element_text(angle = 30, hjust = 1))
```

## Twin test

The first and the second crawl are the same movement. An 8-label model that separates them is
learning the position in the protocol (fatigue, order) rather than the movement, and that would
also inflate any 8-class result.

```{r twin-test}
if (length(twin_results)) {
  tw <- merge(rbindlist(twin_results), seg_meta[, .(seg_id, class, seg_label)], by = "seg_id")
  crawl <- tw[class == crawl_class]
  print(crawl[, table(true_segment = seg_label, predicted = pred)])
  cat(sprintf("crawl segments recognised as crawling: %.1f%%; first vs second crawl told apart: %.1f%% (50%% = indistinguishable)\n",
              100 * crawl[, mean(startsWith(pred, paste0(crawl_class, "_")))], 100 * crawl[, mean(pred == seg_label)]))
}
```

## Feature importance (last fold, all-feature XGBoost)

```{r importance}
imp <- xgboost::xgb.importance(model = xg$model)
knitr::kable(head(imp, 25), digits = 3)
```

# 7. SOM maps

These maps are descriptive: they are trained on all children in the run (no held-out data),
to look at what the maps learn.

## 7.1 DTW-SOM on whole segments

Every segment enters the map at its own length. A prototype keeps the length of the segment it
was initialised from, and batch training averages the mapped segments along their DTW paths
(DBA-style), so the codebooks end up with different lengths.

```{r map-dtwsom, eval = run_maps}
sc_map <- fit_scaler(seg_lo)
w_map <- max(2L, round(0.1 * median(vapply(seg_lo, nrow, 1L))))
map_fit <- dtwsom::dtw_som(apply_scaler_list(seg_lo, sc_map), xdim = params$som_xdim, ydim = params$som_ydim,
                           topo = "hexagonal", epochs = if (params$quick) 20 else 40, mode = "batch",
                           window = w_map, smooth = 3, normalize = "none", threads = n_cores,
                           seed = params$seed, verbose = FALSE)
print(map_fit)
map_labels <- seg_meta$class
ch <- match(params$plot_channel, colnames(seg_lo[[1]])); if (is.na(ch)) ch <- 1L
K_map <- length(map_fit$codes)
tab_map <- table(unit = factor(map_fit$unit.classif, levels = seq_len(K_map)), task = factor(map_labels, levels = classes))
unit_info <- data.frame(unit = seq_len(K_map), n = rowSums(tab_map),
                        majority = ifelse(rowSums(tab_map) > 0, classes[max.col(tab_map, ties.method = "first")], NA),
                        purity = round(ifelse(rowSums(tab_map) > 0, apply(tab_map, 1, max) / pmax(rowSums(tab_map), 1), NA), 2),
                        codebook_s = vapply(map_fit$codes, nrow, 1L) / params$fs_seg)
cat(sprintf("Sakoe-Chiba window %d samples; ARI(units, task) = %.3f\n", w_map, ari(map_labels, map_fit$unit.classif)))
knitr::kable(unit_info, caption = "Units: hits, majority task, purity, codebook length in seconds")
```

**The map.** Each hexagon is a unit, filled with its majority task (stronger colour = purer unit).
The black line is the codebook (grey for units that no segment chose), the thin lines are the
segments mapped to it, coloured by their task. All lines share one time scale (bar at the bottom), so a short codebook stops early and the
different lengths are visible directly. The label gives the unit number, hits and codebook length.

```{r hex-map-function, eval = run_maps}
hex_series_map <- function(fit, labels, channel = 1, sc = NULL, fs = 10, pal, max_members = 20,
                           title = NULL, subtitle = NULL, seed = 1) {
  g <- fit$grid; K <- nrow(g$pts); lv <- names(pal)
  unscale <- function(v) if (is.null(sc)) v else v * sc$sd[channel] + sc$mu[channel]
  bmu <- fit$unit.classif
  tab <- table(factor(bmu, levels = seq_len(K)), factor(labels, levels = lv))
  n_u <- rowSums(tab)
  major <- ifelse(n_u > 0, lv[max.col(tab, ties.method = "first")], NA_character_)
  purity <- ifelse(n_u > 0, apply(tab, 1, max) / pmax(n_u, 1), 0)

  ang <- seq(30, 330, by = 60) * pi / 180; R <- 0.5 / cos(pi / 6)       # pointy-top hexagons, like kohonen
  hex <- data.frame(unit = rep(seq_len(K), each = 6),
                    x = rep(g$pts[, 1], each = 6) + R * cos(ang),
                    y = rep(g$pts[, 2], each = 6) + R * sin(ang))
  hex$major <- factor(major[hex$unit], levels = lv)
  hex$purity <- purity[hex$unit]

  set.seed(seed)
  shown <- unlist(lapply(seq_len(K), function(k) { m <- which(bmu == k); if (length(m) > max_members) sample(m, max_members) else m }))
  Tmax <- max(vapply(fit$codes, nrow, 1L), vapply(fit$data[shown], nrow, 1L))
  vals <- unscale(c(unlist(lapply(fit$codes, function(m) m[, channel])), unlist(lapply(fit$data[shown], function(m) m[, channel]))))
  vr <- quantile(vals, c(0.01, 0.99), names = FALSE)
  box_w <- 0.84; y_lo <- -0.30; y_hi <- 0.20                          # series box inside the hexagon
  to_xy <- function(v, k) {
    v <- unscale(v)
    data.frame(x = g$pts[k, 1] - box_w / 2 + box_w * (seq_along(v) - 1) / (Tmax - 1),
               y = g$pts[k, 2] + y_lo + (y_hi - y_lo) * pmin(pmax((v - vr[1]) / diff(vr), 0), 1))
  }
  mem <- do.call(rbind, lapply(shown, function(s) cbind(to_xy(fit$data[[s]][, channel], bmu[s]), id = s,
                                                         class = factor(labels[s], levels = lv))))
  cod <- do.call(rbind, lapply(seq_len(K), function(k) cbind(to_xy(fit$codes[[k]][, channel], k), unit = k,
                                                              used = n_u[k] > 0)))
  txt <- data.frame(x = g$pts[, 1], y = g$pts[, 2] + 0.36,
                    label = sprintf("%d: n=%d, %.0fs", seq_len(K), n_u, vapply(fit$codes, nrow, 1L) / fs))
  bar_w <- box_w * (10 * fs) / (Tmax - 1)
  x0 <- min(g$pts[, 1]) - 0.42; y0 <- min(g$pts[, 2]) - 0.72
  ggplot() +
    geom_polygon(data = hex, aes(x, y, group = unit, fill = major, alpha = purity), colour = "white", linewidth = 0.8) +
    geom_path(data = mem, aes(x, y, group = id, colour = class), linewidth = 0.3, alpha = 0.6) +
    geom_path(data = cod[!cod$used, ], aes(x, y, group = unit), colour = "grey55", linewidth = 0.45) +
    geom_path(data = cod[cod$used, ], aes(x, y, group = unit), colour = "black", linewidth = 0.6) +
    geom_text(data = txt, aes(x, y, label = label), size = 2.3, colour = "grey15") +
    annotate("segment", x = x0, xend = x0 + bar_w, y = y0, yend = y0, linewidth = 0.8) +
    annotate("text", x = x0 + bar_w + 0.06, y = y0, label = "10 s", hjust = 0, size = 3) +
    scale_fill_manual(values = pal, na.value = "grey92", drop = FALSE, name = "task") +
    scale_colour_manual(values = pal, guide = "none") +
    scale_alpha(range = c(0.12, 0.5), guide = "none") +
    guides(fill = guide_legend(override.aes = list(alpha = 0.5))) +
    coord_equal() + theme_void(base_size = 11) +
    labs(title = title, subtitle = subtitle)
}
```

```{r map-hex, eval = run_maps, fig.width = 10, fig.height = 9}
hex_series_map(map_fit, map_labels, channel = ch, sc = sc_map, fs = params$fs_seg, pal = pal,
               title = "DTW-SOM on whole segments: codebooks (black) and mapped segments",
               subtitle = sprintf("channel %s, %d Hz; up to 20 segments drawn per unit", colnames(seg_lo[[1]])[ch], params$fs_seg))
```

**Lengths.** Each unit's codebook length (black diamond) next to the durations of the segments
mapped to it. A good map puts segments near their codebook's length; DTW absorbs the rest.

```{r map-lengths, eval = run_maps, fig.height = 4.5}
mem_len <- data.frame(unit = factor(map_fit$unit.classif, levels = seq_len(K_map)),
                      dur_s = vapply(map_fit$data, nrow, 1L) / params$fs_seg, task = map_labels)
ggplot() +
  geom_jitter(data = mem_len, aes(unit, dur_s, colour = task), width = 0.2, height = 0, alpha = 0.75, size = 1.4) +
  geom_point(data = unit_info, aes(factor(unit, levels = seq_len(K_map)), codebook_s, fill = n > 0), shape = 23, size = 3.2) +
  scale_fill_manual(values = c(`TRUE` = "black", `FALSE` = "white"), guide = "none") +
  scale_colour_manual(values = pal) + scale_x_discrete(drop = FALSE) +
  labs(x = "unit", y = "seconds", title = "Codebook length (diamond) and durations of the mapped segments") +
  theme_minimal()
```

**The package's own views.** `plot(fit, "members")` draws every unit with its members on their
own sample index; `"members_dtw"` warps each member onto its codebook's time axis along the DTW
path, which shows how well the codebook describes its members.

```{r map-members, eval = run_maps, fig.width = 11, fig.height = 9}
plot(map_fit, "members", channel = ch)
```

```{r map-members-dtw, eval = run_maps, fig.width = 11, fig.height = 9}
plot(map_fit, "members_dtw", channel = ch)
```

```{r map-quality, eval = run_maps, fig.width = 12, fig.height = 4.5}
op <- par(mfrow = c(1, if (has("MASS")) 4 else 3), mar = c(1, 1, 3, 1))
plot(map_fit, "counts", main = "hits per unit")
plot(map_fit, "umatrix", main = "U-matrix (DTW)")
if (has("MASS")) plot(map_fit, "elastic", main = "elastic layout")
par(mar = c(4, 4, 3, 1)); plot(map_fit, "qe")
par(op)
```

## 7.2 Feature SOM (the article's view)

Segment-level features (mean over the segment's windows), standardised, on a 5x5 hexagonal
kohonen map. Background colour is the unit's majority task; thick lines are the boundaries of
the 7 super-clusters (Ward on the codebook vectors).

```{r map-kohonen, fig.width = 11, fig.height = 6}
sdv <- apply(feat_seg, 2, sd, na.rm = TRUE)
Zs <- scale(feat_seg[, is.finite(sdv) & sdv > 0, drop = FALSE]); Zs[!is.finite(Zs)] <- 0
set.seed(params$seed)
koh <- kohonen::som(Zs, grid = kohonen::somgrid(params$som_xdim, params$som_ydim, "hexagonal"), rlen = 500, mode = "batch")
K_koh <- nrow(koh$codes[[1]])
tab_koh <- table(factor(koh$unit.classif, levels = seq_len(K_koh)), factor(seg_meta$class, levels = classes))
major_koh <- ifelse(rowSums(tab_koh) > 0, classes[max.col(tab_koh, ties.method = "first")], NA)
bg_koh <- rep("grey95", K_koh); bg_koh[!is.na(major_koh)] <- adjustcolor(pal[major_koh[!is.na(major_koh)]], 0.3)
super_koh <- super_clusters(as.matrix(dist(koh$codes[[1]])), length(classes))
cat(sprintf("%d segments x %d features; ARI(units, task) = %.3f; ARI(super-clusters, task) = %.3f\n",
            nrow(Zs), ncol(Zs), ari(seg_meta$class, koh$unit.classif), ari(seg_meta$class, super_koh[koh$unit.classif])))

layout(matrix(c(1, 2, 3, 3), 2, byrow = TRUE), heights = c(6, 1))
op <- par(mar = c(0.5, 0.5, 2.5, 0.5))
plot(koh, type = "mapping", pchs = 19, col = adjustcolor(pal[seg_meta$class], 0.85), bgcol = bg_koh,
     shape = "straight", main = "Segments on the feature SOM")
kohonen::add.cluster.boundaries(koh, super_koh, lwd = 3)
plot(koh, type = "dist.neighbours", shape = "straight", main = "U-matrix (feature space)",
     palette.name = function(n) hcl.colors(n, "Inferno", rev = TRUE))
par(mar = c(0, 0, 0, 0)); plot.new()
legend("center", legend = classes, pch = 19, col = pal, horiz = TRUE, bty = "n", cex = 0.95)
par(op); layout(1)
```

**Component planes** of the most important XGBoost features: where on the map each feature is high.

```{r map-kohonen-planes, fig.width = 11, fig.height = 7}
top_f <- head(intersect(imp$Feature, colnames(Zs)), 6)
op <- par(mfrow = c(2, 3), mar = c(0.5, 0.5, 2.5, 0.5))
for (fn in top_f) plot(koh, type = "property", property = koh$codes[[1]][, fn], main = fn, shape = "straight",
                       palette.name = function(n) hcl.colors(n, "viridis"))
par(op)
```

```{r map-kohonen-training, fig.width = 10, fig.height = 4.5}
op <- par(mfrow = c(1, 2), mar = c(1, 1, 3, 1))
plot(koh, type = "counts", shape = "straight", main = "hits per unit", palette.name = function(n) hcl.colors(n, "YlGnBu", rev = TRUE))
par(mar = c(4, 4, 3, 1)); plot(koh, type = "changes", main = "training progress")
par(op)
```

# 8. Next steps

* Run with `quick: false`, `k_folds: 9`, `repeats: 5` for the full protocol (hours with DTW-SOM and Mantis on).
* Learning curve: repeat the loop with 9, 18, 36, 72 training children and plot accuracy per method.
* Embedding fusion: concatenate Mantis embeddings with the RQA + catch22 matrix before the random forest.
* `chaosrcpp`: automatic `tau` (average mutual information) and `m` (false nearest neighbours) for the
  magnitude channels, cross-RQA between accelerometer and gyroscope, the 0-1 test for chaos.
* `dtwsom`: a `map_units_to_labels()` helper (the logic of `unit_to_class()` above), an LB_Keogh
  lower bound, and a class-coloured variant of `plot_members()`.

```{r session}
sessionInfo()
```
