---
title: "A robust Self-Organizing Map for children's movement tasks"
subtitle: "An improvement of the SOM in Patrikei et al. (2026), *Acta Gymnica*: the same features and pipeline, a more accurate map"
author: "Andrii Patrikei"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  html_document:
    toc: true
    toc_float:
      collapsed: false
    toc_depth: 2
    code_folding: hide
    theme: flatly
    highlight: tango
params:
  article_rds: "data/som_gain_2_35f_5x5.RDS"   # the article's saved maps; downloaded from its repository if missing
  quick: false                                 # TRUE = a small plan (few seeds and maps) to test the document in minutes
  cores: 0                                     # 0 = all but two; used only when results/ lacks a part of the plan
  save_figures: true                           # also write the static PNGs and the results table used by the README
---

```{css, echo = FALSE}
.main-container { max-width: 1120px; }
body { font-size: 16px; line-height: 1.62; color: #1d1d1b; }
h1.title { font-size: 34px; margin-top: 18px; }
h3.subtitle { font-size: 19px; color: #52514e; line-height: 1.4; }
h1 { font-size: 26px; margin-top: 44px; } h2 { font-size: 20px; margin-top: 30px; }
.lead { font-size: 18px; line-height: 1.6; }
.tiles { display: grid; grid-template-columns: repeat(auto-fit, minmax(200px, 1fr)); gap: 14px; margin: 22px 0 10px; }
.tile { border: 1px solid #e6e5e0; border-left: 4px solid #eb6834; border-radius: 6px; padding: 12px 16px; background: #fcfcfb; }
.tile .num { font-size: 30px; font-weight: 700; line-height: 1.15; color: #0b0b0b; }
.tile .lbl { font-size: 14px; color: #1d1d1b; margin-top: 4px; }
.tile .sub { font-size: 13px; color: #52514e; margin-top: 2px; }
.callout { border-left: 4px solid #2a78d6; background: #f3f7fd; padding: 12px 18px; margin: 18px 0 22px; border-radius: 0 6px 6px 0; }
.callout p:last-child, .callout ul:last-child { margin-bottom: 0; }
.note { color: #52514e; font-size: 14px; }
table { font-size: 14px; }
.table > thead > tr > th, .table > tbody > tr > td { padding: 5px 9px; }
details summary { cursor: pointer; color: #2a78d6; margin: 8px 0; }
pre code { font-size: 13px; }
span.math.display { display: block; overflow-x: auto; overflow-y: hidden; }
```

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, comment = "#>", fig.align = "center", dpi = 110)
library(data.table)
library(ggplot2)
for (f in c("R/article_metrics.R", "R/som_fits.R", "R/protocols.R", "R/bench.R", "R/plots.R", "R/plots_interactive.R")) source(f)
load_novel_som()
cores <- if (params$cores > 0) params$cores else max(1L, parallel::detectCores() - 2L)
res_dir <- if (params$quick) "results_quick" else "results"
```

```{r load, include = FALSE}
# the article's saved maps and feature matrices (downloaded once, checked by md5), the plan, and the results
D <- article_data(get_article_rds(params$article_rds))
plan <- if (!params$quick) bench_plan() else {
  q <- c("kohonen", "euclid:article", "l1_huber:article", "huber:article", "euclid:novel", "huber:novel")
  list(main_all = make_jobs(names(FEATURE_SETS), 27, c(q, "sbd:article", "dtw:novel", "correlation:article"), 1:3, "all"),
       main_cv = make_jobs(names(FEATURE_SETS), 27, q, 1:3, c("kfold", "heldout")),
       curve = make_jobs(names(FEATURE_SETS), c(10, 27), c("kohonen", "l1_huber:article", "euclid:novel"), 1:2, "all"),
       confirm = make_jobs(names(FEATURE_SETS), 27, c("kohonen", "l1_huber:article"), 101:102, "preds"))
}
res <- compute_all(D, plan, cores, dir = res_dir)
for (p in names(plan)) if (!is.null(attr(res[[p]], "failed"))) stop("failed jobs in ", p)
stopifnot(nrow(res$main_all) == nrow(plan$main_all), nrow(res$main_cv) == nrow(plan$main_cv),
          nrow(res$curve) == nrow(plan$curve), nrow(res$confirm) == 648 * nrow(plan$confirm))
```

```{r analysis, include = FALSE}
SETS <- FEATURE_SETS
CV_COLS <- c("kfold_train", "kfold_test", "heldout_super_acc", "heldout_super_bacc", "heldout_unit_acc", "heldout_unit_bacc")
R <- merge(res$main_all, res$main_cv[, c("set", "n", "method", "seed", CV_COLS), with = FALSE],
           by = c("set", "n", "method", "seed"), all.x = TRUE)
MEASURES <- c("QE", "EV", "TE", "KL", "accuracy", "mnLogLoss", "accuracy_euclid", "ari_super", CV_COLS, "seconds")
long <- melt(R, id.vars = c("set", "n", "method", "seed"), measure.vars = MEASURES, variable.name = "measure")[!is.na(value)]
S <- long[, .(mean = mean(value), sd = sd(value), min = min(value), max = max(value), seeds = .N), by = .(set, n, method, measure)]
S <- data.table(S, method_parts(S$method))
S[, set_label := factor(SETS[set], levels = SETS)]
pubs <- as.data.table(D$published)[, set_label := factor(SETS[set], levels = SETS)]

# the screening rule, fixed before looking at the results: among the variants whose distance means something on a
# feature vector, trained with the article's schedule, the best held-out accuracy (super-cluster rule) averaged
# over the five feature sets
overall <- dcast(S[n == 27, .(mean = mean(mean)), by = .(method, label, regime, elastic, measure)],
                 method + label + regime + elastic ~ measure, value.var = "mean")
best <- overall[regime == "article" & !elastic & !is.na(heldout_super_acc)][order(-heldout_super_acc)]$method[1]
best_own <- overall[regime == "novel" & !elastic & !is.na(heldout_super_acc)][order(-heldout_super_acc)]$method[1]
confirmed <- setdiff(unique(plan$confirm$method), "kohonen")
if (!identical(best, confirmed)) warning("the screening picks ", best, " but the confirmatory run tests ", confirmed)
CLASSIC <- "kohonen"; ROBUST <- best
kh <- overall[method == CLASSIC]; rb <- overall[method == ROBUST]; ow <- overall[method == best_own]
per_set <- function(me, m, nn = 27) setNames(S[n == nn & measure == me & method == m][match(names(SETS), set)]$mean, names(SETS))
f3 <- function(x) sprintf("%.3f", x)
pts <- function(x) sub("^[-+]0[.]0$", "0.0", sprintf("%+.1f", 100 * x))   # percentage points, no sign on a zero

# robust minus classical, per feature set and measure, with a 95% Welch interval over the seeds
DD <- rbindlist(lapply(c("heldout_super_acc", "accuracy", "heldout_unit_acc", "kfold_test", "heldout_super_bacc"), function(me)
  rbindlist(lapply(names(SETS), function(s) {
    v <- R[n == 27 & set == s & method == ROBUST & !is.na(get(me))][[me]]; k <- R[n == 27 & set == s & method == CLASSIC & !is.na(get(me))][[me]]
    ci <- if (length(v) > 2 && sd(c(v, k)) > 0) t.test(v, k)$conf.int else c(NA, NA)
    data.table(set = s, measure = me, robust = mean(v), classic = mean(k), diff = mean(v) - mean(k), lo = ci[1], hi = ci[2], seeds = length(v))
  }))))
DD[, set_label := factor(SETS[set], levels = SETS)]
d_ho <- DD[measure == "heldout_super_acc"]; d_all <- DD[measure == "accuracy"]
sets_ho <- d_ho[!is.na(lo) & lo > 0]$set; sets_all <- d_all[!is.na(lo) & lo > 0]$set
level_ho <- setdiff(names(SETS), sets_ho)
and_list <- function(x) if (length(x) <= 1) paste(x, collapse = "") else paste(paste(x[-length(x)], collapse = ", "), "and", x[length(x)])
ho_cl <- per_set("heldout_super_acc", CLASSIC); ho_rb <- per_set("heldout_super_acc", ROBUST)
spec_beats_all <- ho_rb[["spectr"]] > max(ho_cl)

# child by child, on the fresh seeds of the confirmatory run
CF <- copy(res$confirm)[, `:=`(ok_super = pred_super == y7, ok_unit = pred_unit == y7)]
seed_acc <- CF[, .(super = mean(ok_super), unit = mean(ok_unit)), by = .(set, method, seed)]
child_acc <- CF[, .(super = mean(ok_super), unit = mean(ok_unit)), by = .(set, method, child)]
child_test <- function(classic, robust) {
  d <- robust - classic; set.seed(1)
  boot <- replicate(10000, mean(sample(d, replace = TRUE)))
  p <- if (any(d != 0)) suppressWarnings(wilcox.test(robust, classic, paired = TRUE, exact = FALSE))$p.value else 1
  data.table(better = sum(d > 0), worse = sum(d < 0), equal = sum(d == 0), mean_gain = mean(d),
             lo = unname(quantile(boot, 0.025)), hi = unname(quantile(boot, 0.975)), p = p)
}
CT <- rbindlist(lapply(c("super", "unit"), function(rule) rbindlist(lapply(names(SETS), function(s) {
  x <- dcast(child_acc[set == s], child ~ method, value.var = rule)
  cbind(set = s, rule = rule, child_test(x[[CLASSIC]], x[[ROBUST]]))
}))))
CT[, p_holm := p.adjust(p, "holm"), by = rule]
CT[, set_label := factor(SETS[set], levels = SETS)]
PC <- dcast(child_acc, set + child ~ method, value.var = "super")
setnames(PC, c(CLASSIC, ROBUST), c("classic", "robust"))
PC[, set_label := factor(SETS[set], levels = SETS)]
n_children <- length(unique(D$child))
ct_spec <- CT[rule == "super" & set == "spectr"]; ct_ts <- CT[rule == "super" & set == "tsfresh"]
rep_acc <- seed_acc[, .(super = mean(super), unit = mean(unit)), by = .(set, method)]
n_all <- res$main_all[method == CLASSIC, uniqueN(seed)]; n_cv <- res$main_cv[method == CLASSIC, uniqueN(seed)]
n_conf <- res$confirm[method == CLASSIC, uniqueN(seed)]; conf_seeds <- range(res$confirm$seed)
# the article's way of reading the robust codebook (Euclidean winners, PAM): how often it still beats kohonen
eu_rb <- per_set("accuracy_euclid", ROBUST); eu_cl <- per_set("accuracy_euclid", CLASSIC)
```

<p class="lead">
The article clustered 648 movement segments (81 children, 7 tasks, crawling recorded twice) with a classical
Kohonen Self-Organizing Map and judged how well its clusters recover the tasks. Here the map alone is
replaced by a **robust SOM** from [novel_SOM](https://github.com/andrii-patrikei/novel_SOM): the winning unit is
chosen by the Manhattan distance, every update step is clipped at one standard deviation, and the training
schedule is exactly the article's. The features, the 5&times;5 map, the 7 super-clusters and the accuracy measure
are the article's own, read from its saved models.
</p>

<div class="tiles">
<div class="tile"><div class="num">`r pts(rb$heldout_super_acc - kh$heldout_super_acc)` points</div>
<div class="lbl">accuracy on children the map has never seen, mean of the 5 feature sets</div>
<div class="sub">from `r f3(kh$heldout_super_acc)` to `r f3(rb$heldout_super_acc)`</div></div>
<div class="tile"><div class="num">`r pts(per_set("heldout_super_acc", ROBUST)[["spectr"]] - per_set("heldout_super_acc", CLASSIC)[["spectr"]])` points</div>
<div class="lbl">on the spectral features, the largest gain</div>
<div class="sub">from `r f3(ho_cl[["spectr"]])` to `r f3(ho_rb[["spectr"]])``r if (spec_beats_all) ", above the classical map on any feature set" else ""`</div></div>
<div class="tile"><div class="num">`r ct_spec$better` of `r n_children`</div>
<div class="lbl">children recognised better on the spectral features</div>
<div class="sub">worse for `r ct_spec$worse`; checked on fresh seeds, child by child</div></div>
<div class="tile"><div class="num">`r length(sets_ho)` of 5</div>
<div class="lbl">feature sets with a gain beyond the seed noise on held-out children</div>
<div class="sub">95% intervals exclude zero`r if (length(level_ho)) paste0("; ", and_list(SETS[level_ho]), " level") else ""`</div></div>
</div>

```{r headline, echo = FALSE}
S2 <- S[n == 27 & method %in% c(CLASSIC, ROBUST) & measure %in% c("heldout_super_acc", "accuracy")]
S2[, who := ifelse(method == CLASSIC, "classic", "robust")][, protocol := as.character(measure)]
p_dumbbell(as.data.frame(S2), c("heldout_super_acc", "accuracy"),
           c(sprintf("Children never seen (%d seeds)", max(S2[protocol == "heldout_super_acc"]$seeds)),
             sprintf("All 648 segments (%d seeds)", max(S2[protocol == "accuracy"]$seeds))))
```

<p class="note">Each dot is a mean over seeds; hover for the spread. The number beside each pair is the gain in
percentage points. The buttons switch between children the map has never seen and the article's own
accuracy on all 648 segments.</p>

What changes in the code is one call:

```{.r .fold-show}
# the article (Patrikei et al. 2026, functions/functions_df_total_cv.R)
kohonen::som(X, grid = kohonen::somgrid(5, 5, "hexagonal"), rlen = 1000, alpha = c(0.05, 0.01),
             dist.fcts = "sumofsquares", init = aweSOM::somInit(X, 5, 5, "random"), maxNA.fraction = 8)
# this report: the novel_SOM core with an L1 winner and a clipped pull, on the same schedule
fit_som(X, "l1_huber:article", seed)          # R/som_fits.R
```

# 1. What is compared

## The article's pipeline

1. **Data.** 81 children, eight segments each recorded with a back-mounted IMU (Cuberek et al., 2024): a 4x10 m
   shuttle run, crawling, a progressive cone shuffle run, a ring collection and placement run, a 10x position
   change drill, 10x gymnastic hoops passing, 10x bench jumping, and crawling again. Seven tasks, 648 segments.
2. **Features.** Five feature sets (RQA; autocorrelation; spectral; Mix, which is the three together; all tsfresh
   features), each reduced to its top 27 features by XGBoost gain and standardised.
3. **Map.** A 5&times;5 hexagonal SOM, `kohonen::som`, online, 1000 passes, learning rate 0.05 to 0.01.
4. **Clusters and accuracy.** The 25 prototypes are cut into 7 super-clusters by PAM; each super-cluster is named
   after the task most of its segments belong to, and the accuracy is the share of segments whose super-cluster
   carries their task.

## The robust SOM

The classical online SOM picks the unit nearest to a segment $\mathbf{x}$ in Euclidean distance and pulls the
winner and its lattice neighbours toward $\mathbf{x}$. The robust SOM changes two things and nothing else:

$$c(\mathbf{x}) = \arg\min_k \sum_{i=1}^{27} \lvert x_i - m_{k,i} \rvert$$

$$\mathbf{m}_j \leftarrow \mathbf{m}_j + \alpha(t)\, h_{c j}(t)\, \operatorname{clip}\!\left(\mathbf{x} - \mathbf{m}_j,\, -1,\, 1\right)$$

* **The winner** is chosen by the Manhattan (L1) distance, which a single extreme feature cannot dominate.
* **The pull** is clipped at one standard deviation in every feature, so an outlying segment moves a prototype
  no further than a typical one.
* **Everything else is the article's:** the bubble neighbourhood $h_{cj}(t) = 1$ when the lattice distance between
  units $c$ and $j$ is at most the radius $r(t)$; $\alpha$ from 0.05 to 0.01 and $r$ from 3 to 0 over 1000
  passes; the start, 25 randomly drawn segments. The super-clusters are PAM on the L1 distances between the
  prototypes.

The variant comes from [novel_SOM](https://github.com/andrii-patrikei/novel_SOM), a compiled SOM core with
seventeen variants: nine distances, six with a training ingredient (a clipped pull, a conscience, annealing or
momentum) and two lattices (a rectangular sheet and a torus). Its code is vendored in `novel_som/`, with one
addition: the bubble neighbourhood of kohonen, which lets any variant run on the article's schedule.

## How it is judged

| | Held fixed (the article's) | Changed |
|---|---|---|
| Data | the article's feature matrices, read from its saved models | |
| Map | 5&times;5 units, 1000 passes, learning rate 0.05 to 0.01 | the SOM algorithm |
| Clusters | 7 super-clusters by PAM, each named after its majority task | PAM uses the map's own distance |
| Measures | accuracy, QE, explained variance, topographic and Kaski-Lagus error, the article's k-fold | plus children held out |
| Seeds | | `r n_all` seeds per map instead of one |

Three protocols, all with 27 features:

* **All data**, the article's Table 4: one map on all 648 segments (`r n_all` seeds).
* **Held-out children**: 9 folds of 9 children; the map, its super-clusters and their names come from 72
  children, and the segments of the other 9 are predicted (`r n_cv` seeds). This is the honest test of recognising
  movement in a new child.
* **Confirmation**: the screening winner against the classical SOM again, on `r n_conf` fresh seeds (new starts, new
  orders of segments, new folds of children), with every prediction kept for a child-by-child comparison.

# 2. The article, reproduced

The ports of the article's measures (`R/article_metrics.R`, without caret) recompute every number saved with the
article's 170 maps (2 to 35 features, five feature sets):

```{r reproduce, echo = FALSE}
chk <- rbindlist(lapply(2:35, function(n) rbindlist(lapply(seq_along(SETS), function(k) {
  s <- D$models[[n]]$som[[names(SETS)[k]]]$SOM_results$som_model
  q <- as.data.frame(D$models[[n]]$quality[[1]])[k, ]
  a <- article_accuracy(s$unit.classif, cluster::pam(s$codes[[1]], 7)$clustering, D$y7)
  sq <- som_quality(s$data[[1]], s$codes[[1]], kohonen::unit.distances(s$grid))
  data.table(Accuracy = a$accuracy - q[["Accuracy rate all"]], mnLogLoss = a$mnLogLoss - unname(q[["mnLogLoss all"]]),
             `Quantization error` = sq$QE - q[["Quantization error"]], `Explained variance` = sq$EV - q[["(% explained variance)"]],
             `Topographic error` = sq$TE - q[["Topographic error"]], `Kaski-Lagus error` = sq$KL - q[["Kaski-Lagus error"]])
}))))
knitr::kable(data.frame(Measure = names(chk), `Largest difference over the 170 saved maps` = signif(sapply(chk, function(v) max(abs(v))), 2),
                        check.names = FALSE), row.names = FALSE)
pub <- D$models[[27]]$som$tsfresh$SOM_results$som_model
re <- fit_kohonen(D$X$tsfresh[["27"]], 1245)
same <- rbindlist(lapply(1:10, function(s) {
  k <- fit_kohonen(D$X$tsfresh[["27"]], s); e <- fit_novel(D$X$tsfresh[["27"]], "euclid", s, regime = "article")
  data.table(seed = s, identical = max(abs(k$M - e$M)) == 0)
}))
```

The article's published map also comes back bit for bit: `set.seed(1245)` and its own `kohonen::som` call give
the published codebook again (largest difference `r max(abs(re$M - pub$codes[[1]]))`). And the novel_SOM core on
the article's schedule *is* kohonen's algorithm: with the same seed it draws the same start and the same order of
segments and returns an identical codebook in `r sum(same$identical)` of 10 seeds. In the others the two runs part
at the first near tie in the winner search, which kohonen settles with an extra random draw (distances within a
relative 10<sup>-8</sup>) and novel_SOM by taking the first unit; from there on they are two equally valid runs of
the same algorithm. A single run is therefore partly luck, and every map below is trained with `r n_all` seeds.

# 3. Screening 34 SOMs

Every novel_SOM variant was trained twice: with its own schedule (Gaussian neighbourhood, learning rate 0.5 to
0.01, 10 passes) and with the article's. Pick a measure; rows are sorted by the held-out accuracy (empty cells:
not judged on held-out children). "&middot; article's" marks the article's schedule, "&middot; own" novel_SOM's own.

```{r leaderboard, echo = FALSE}
sched <- function(m) ifelse(m == "kohonen", "", ifelse(grepl(":article$", m), " (article's schedule)", " (own schedule)"))
lab_of <- function(m) ifelse(m == CLASSIC, "Classical SOM, kohonen::som (the article)",
                             ifelse(m == ROBUST, "Robust SOM: Manhattan winner, clipped pull (article's schedule)", paste0(method_parts(m)$label, sched(m))))
meas <- c(heldout_super_acc = "Held-out children, the article's super-cluster rule", heldout_unit_acc = "Held-out children, each unit names its task",
          accuracy = "All 648 segments, the article's accuracy", kfold_test = "The article's k-fold test", TE = "Topographic error (lower is better)")
ordr <- overall[order(-heldout_super_acc, -accuracy)]$method
mats <- lapply(names(meas), function(me) {
  x <- dcast(S[n == 27 & measure == me], method ~ set, value.var = "mean")
  m <- as.matrix(x[match(ordr, method), names(SETS), with = FALSE]); cbind(m, `Mean of the five` = rowMeans(m))
})
short_of <- function(m) ifelse(m == CLASSIC, "Classical SOM (the article)", ifelse(m == ROBUST, "Robust SOM (L1 winner, clipped pull)",
                               paste0(method_parts(m)$label, ifelse(grepl(":article$", m), " \u00b7 article's", " \u00b7 own"))))
p_leaderboard(mats, short_of(ordr), c("RQA", "Autocorr.", "Spectral", "Mix", "tsfresh", "Mean"), unname(meas), names(meas) == "TE",
              bold = match(c(CLASSIC, ROBUST), ordr))
```

<div class="callout">

* **On its own schedule, novel_SOM gains little.** Its maps are better ordered (topographic error
  `r sprintf("%.2f", ow$TE)` for its best variant, against `r sprintf("%.2f", kh$TE)`), but that variant (`r ow$label`)
  reaches `r f3(ow$heldout_super_acc)` on held-out children against `r f3(kh$heldout_super_acc)` for the classical SOM
  (`r pts(ow$heldout_super_acc - kh$heldout_super_acc)` points).
* **On the article's schedule, the robust variants lead.** The Euclidean variant on this schedule matches the
  classical SOM (held-out `r f3(overall[method == "euclid:article"]$heldout_super_acc)` against `r f3(kh$heldout_super_acc)`), as
  it should, since it is the same algorithm. What lifts the accuracy is robustness: an L1 winner, a clipped pull,
  a conscience against units that collect the extreme segments.
* **The winner by the rule fixed in advance** (best held-out accuracy over the five feature sets, among the
  variants whose distance means something on a feature vector) is the **Manhattan winner with the clipped
  pull**: `r f3(rb$heldout_super_acc)` on held-out children.
* The elastic distances (DTW, soft-DTW, shape-based) align a series in time, and a feature vector has no time
  axis: on all data the DTW family reaches at most `r f3(max(overall[method %in% c("dtw:article", "dtw:novel", "softdtw:novel", "dtw_anneal:article", "dtw_anneal:novel")]$accuracy))`
  and the shape-based distance `r f3(max(overall[method %in% c("sbd:article", "sbd:novel")]$accuracy))`, against
  `r f3(kh$accuracy)` for the classical SOM.

</div>

<details><summary>All 34 maps in a table (sortable, searchable, downloadable)</summary>

```{r leaderboard-table, echo = FALSE}
lt <- overall[match(ordr, method), .(Map = lab_of(method), Schedule = c(kohonen = "the article's (kohonen)", article = "the article's", novel = "novel_SOM's own")[regime],
                                     `Elastic distance` = ifelse(elastic, "yes", ""), `Held-out, super-clusters` = heldout_super_acc,
                                     `Held-out, unit majority` = heldout_unit_acc, `All data (article)` = accuracy, `Article's k-fold test` = kfold_test,
                                     `Topographic error` = TE)]
ho_sets <- dcast(S[n == 27 & measure == "heldout_super_acc"], method ~ set, value.var = "mean")
lt <- cbind(lt, ho_sets[match(ordr, method), names(SETS), with = FALSE]); setnames(lt, names(SETS), paste("Held-out,", SETS))
dt_table(lt, round_cols = names(lt)[4:ncol(lt)], page = 12)
```

</details>

# 4. The robust SOM against the classical SOM

## Seed by seed

```{r seeds, echo = FALSE}
Rs <- melt(R[n == 27 & method %in% c(CLASSIC, ROBUST)], id.vars = c("set", "method", "seed"),
           measure.vars = c("heldout_super_acc", "accuracy"), variable.name = "protocol")[!is.na(value)]
Rs[, who := ifelse(method == CLASSIC, "classic", "robust")][, set_label := factor(SETS[set], levels = SETS)][, protocol := as.character(protocol)]
p_seed_boxes(as.data.frame(Rs), as.data.frame(pubs[n == 27, .(set_label, value = accuracy)]), c("heldout_super_acc", "accuracy"),
             c(sprintf("Children never seen (%d seeds)", n_cv), sprintf("All 648 segments (%d seeds)", n_all)))
```

<p class="note">Every dot is one seed. The black diamonds (all-data view) are the article's published maps:
one seed each, which is why they can sit anywhere in the classical SOM's spread.</p>

## Beyond the seed noise

```{r forest, echo = FALSE}
dd2 <- DD[measure %in% c("heldout_super_acc", "accuracy")]
lv <- c(sprintf("children the map has never seen (%d seeds)", n_cv), sprintf("all 648 segments (%d seeds)", n_all))
dd2[, protocol_label := factor(ifelse(measure == "accuracy", lv[2], lv[1]), levels = lv)]
p_forest(as.data.frame(dd2))
```

```{r forest-table, echo = FALSE}
ft <- dcast(DD[measure %in% c("heldout_super_acc", "accuracy", "heldout_unit_acc")], set_label ~ measure,
            value.var = c("classic", "robust", "diff", "lo", "hi"))
cell <- function(me) sprintf("%s to %s (%s, %s to %s)", f3(ft[[paste0("classic_", me)]]), f3(ft[[paste0("robust_", me)]]),
                            pts(ft[[paste0("diff_", me)]]), pts(ft[[paste0("lo_", me)]]), pts(ft[[paste0("hi_", me)]]))
knitr::kable(data.frame(`Feature set` = ft$set_label, `Held-out children, super-clusters` = cell("heldout_super_acc"),
                        `Held-out children, unit majority` = cell("heldout_unit_acc"), `All 648 segments` = cell("accuracy"), check.names = FALSE),
             caption = "Classical to robust SOM (gain in points, 95% Welch interval over the seeds)")
```

On held-out children the intervals exclude zero on **`r and_list(SETS[sets_ho])`**; on all 648 segments on
**`r and_list(SETS[sets_all])`**. `r if (length(level_ho)) paste0("On ", and_list(SETS[level_ho]), " the gain is within the seed noise", if ("rqa" %in% level_ho) " (RQA is the weakest feature set for every map)" else "", ". ") else ""`The
intervals describe the variation between training runs on the same 81 children; the child-by-child comparison
below tests the children themselves.

## Child by child, on fresh seeds

The screening chose the robust SOM on seeds 1 to `r n_all`. To make sure that choice did not ride on lucky seeds,
the robust and the classical SOM were run again on `r n_conf` new seeds (`r conf_seeds[1]` to `r conf_seeds[2]`: new starts,
new orders of segments, new folds of children), keeping the prediction for every segment. Each child then has an
accuracy under each map, averaged over the `r n_conf` runs and its 8 segments.

```{r children, echo = FALSE}
p_children(as.data.frame(PC), first = SETS[["spectr"]])
```

```{r children-table, echo = FALSE}
ct <- CT[order(rule, set_label)]
knitr::kable(data.frame(`Rule` = ifelse(ct$rule == "super", "super-clusters (the article's)", "unit majority"), `Feature set` = ct$set_label,
                        `Children better / worse / equal` = sprintf("%d / %d / %d", ct$better, ct$worse, ct$equal),
                        `Mean gain per child (points)` = sprintf("%s (%s to %s)", pts(ct$mean_gain), pts(ct$lo), pts(ct$hi)),
                        `Wilcoxon p (Holm)` = ifelse(ct$p_holm < 1e-4, "< 0.0001", sprintf("%.4f", ct$p_holm)), check.names = FALSE),
             caption = sprintf("Held-out children on %d fresh seeds: the gain of each of the %d children (mean, bootstrap 95%% interval over children; paired Wilcoxon signed-rank test, Holm-adjusted over the five feature sets)", n_conf, n_children))
```

With fresh seeds the held-out accuracy averaged over the five feature sets is `r f3(rep_acc[method == CLASSIC, mean(super)])`
for the classical and `r f3(rep_acc[method == ROBUST, mean(super)])` for the robust SOM (`r f3(kh$heldout_super_acc)` and
`r f3(rb$heldout_super_acc)` in the screening), so the gain replicates. On the spectral features
`r ct_spec$better` of the `r n_children` children are recognised better and `r ct_spec$worse` worse; on tsfresh, where the
classical map already gets most children right, `r ct_ts$better` better and `r ct_ts$worse` worse.

## Robustness checks

```{r checks, echo = FALSE}
chk2 <- rbind(
  data.table(Check = "Accuracy, all data (the map's own distance)", Classical = per_set("accuracy", CLASSIC), Robust = per_set("accuracy", ROBUST), set = names(SETS)),
  data.table(Check = "Accuracy, all data, codebook clustered the article's way (Euclidean winners, PAM)", Classical = per_set("accuracy_euclid", CLASSIC),
             Robust = per_set("accuracy_euclid", ROBUST), set = names(SETS)),
  data.table(Check = "Balanced accuracy, held-out children (every task counts the same)", Classical = per_set("heldout_super_bacc", CLASSIC),
             Robust = per_set("heldout_super_bacc", ROBUST), set = names(SETS)),
  data.table(Check = "Topographic error, all data (lower is better)", Classical = per_set("TE", CLASSIC), Robust = per_set("TE", ROBUST), set = names(SETS)))
chk2 <- dcast(chk2, Check ~ set, value.var = c("Classical", "Robust"))
out <- data.frame(Check = chk2$Check)
for (s in names(SETS)) out[[SETS[[s]]]] <- sprintf("%s to %s", f3(chk2[[paste0("Classical_", s)]]), f3(chk2[[paste0("Robust_", s)]]))
knitr::kable(out[match(c("Accuracy, all data (the map's own distance)", "Accuracy, all data, codebook clustered the article's way (Euclidean winners, PAM)",
                         "Balanced accuracy, held-out children (every task counts the same)", "Topographic error, all data (lower is better)"), out$Check), ],
             row.names = FALSE, caption = "Classical to robust SOM, means over the seeds")
```

* **The gain is in the map, not only in the clustering step.** Read the article's way (Euclidean winners, PAM on
  the Euclidean distances between prototypes), the robust codebook still beats the classical SOM on
  `r sum(eu_rb > eu_cl)` of 5 feature sets (mean `r f3(mean(eu_rb))` against `r f3(mean(eu_cl))`; with its own L1 reading
  `r f3(mean(per_set("accuracy", ROBUST)))`).
* **It is not an artefact of class sizes.** Crawling has twice as many segments as the other tasks; the balanced
  accuracy, which weighs every task equally, rises from `r f3(mean(per_set("heldout_super_bacc", CLASSIC)))` to
  `r f3(mean(per_set("heldout_super_bacc", ROBUST)))` on held-out children (mean of the five feature sets).
* **The map stays a map.** Its topographic error averages `r sprintf("%.2f", mean(per_set("TE", ROBUST)))`, against
  `r sprintf("%.2f", mean(per_set("TE", CLASSIC)))` for the classical SOM, so the lattice is `r if (mean(per_set("TE", ROBUST)) <= mean(per_set("TE", CLASSIC))) "at least as well ordered" else "about as well ordered"`.

## The article's Table 4

Every measure of the article's Table 4, for every map and feature set (means over the seeds; the article's
published map, one seed, for comparison). Filter by feature set or map, sort any column.

```{r table4, echo = FALSE}
w <- dcast(S[n == 27], set + method ~ measure, value.var = c("mean", "sd"))
t4 <- w[, .(`Feature set` = SETS[set], Map = lab_of(method), `QE` = mean_QE, `EV %` = mean_EV, TE = mean_TE, KL = mean_KL,
            Accuracy = mean_accuracy, `Accuracy sd` = sd_accuracy, mnLogLoss = mean_mnLogLoss, `k-fold train` = mean_kfold_train,
            `k-fold test` = mean_kfold_test, `Held-out, super-clusters` = mean_heldout_super_acc, `Held-out, units` = mean_heldout_unit_acc)]
p27 <- pubs[n == 27, .(`Feature set` = SETS[set], Map = "The article's published map (seed 1245)", QE, `EV %` = EV, TE, KL, Accuracy = accuracy,
                       `Accuracy sd` = NA_real_, mnLogLoss, `k-fold train` = kfold_train, `k-fold test` = kfold_test,
                       `Held-out, super-clusters` = NA_real_, `Held-out, units` = NA_real_)]
t4 <- rbind(p27, t4)[order(match(`Feature set`, SETS), -Accuracy)]
t4[, `Feature set` := factor(`Feature set`, levels = SETS)]
dt_table(t4, round_cols = names(t4)[3:ncol(t4)], page = 10, filter = "top")
```

<p class="note">QE, EV, TE and KL are the Euclidean measures of <code>aweSOM::somQuality</code>, as in the article.
The article's k-fold "test" trains a new map on the test quarter and scores it on that same quarter, so it is not a
prediction for unseen data; the held-out columns are.</p>

# 5. Why it works: the features have outliers

The variants that win are the robust ones, which points at outliers. The article's features are standardised
but heavy-tailed: a few segments sit far out on a few features, and the squared Euclidean pull lets them drag
prototypes away from the bulk of their task.

```{r tails, echo = FALSE}
tails <- rbindlist(lapply(names(SETS), function(k) {
  X <- D$X[[k]][["27"]]; kurt <- apply(X, 2, function(v) mean(v^4) / mean(v^2)^2)
  data.table(`Feature set` = SETS[[k]], `Segments with a feature beyond 3 sd` = sprintf("%.0f%%", 100 * mean(apply(abs(X) > 3, 1, any))),
             `Largest value (sd)` = sprintf("%.1f", max(abs(X))), `Median kurtosis (Gaussian: 3)` = sprintf("%.1f", median(kurt)),
             `Gain on held-out children` = pts(per_set("heldout_super_acc", ROBUST)[[k]] - per_set("heldout_super_acc", CLASSIC)[[k]]))
}))
knitr::kable(tails, caption = "The 27-feature matrices of the article")
```

A direct test: clip every feature at &plusmn;3 standard deviations (winsorise) and train both maps again. If the
gain is robustness, the classical SOM on clipped features should catch up most of the way, and the robust SOM
should barely change.

```{r winsor, echo = FALSE}
Dw <- D; Dw$models <- NULL
Dw$X <- lapply(D$X, function(s) list(`27` = pmin(pmax(s[["27"]], -3), 3)))
wjobs <- rbind(make_jobs(names(SETS), 27, c(CLASSIC, ROBUST), if (params$quick) 1:2 else 1:20, "all"),
               make_jobs(names(SETS), 27, c(CLASSIC, ROBUST), if (params$quick) 1:2 else 1:10, "heldout"))
W <- cached("somvar_winsor3", bench_key(wjobs[, c("set", "n", "method", "seed", "protocols")], 10, D$md5, "features clipped at 3 sd"),
            function() run_jobs(wjobs, Dw, 10, cores), res_dir)
wsum <- function(x, col) x[!is.na(get(col)), .(v = mean(get(col))), by = .(set, method)]
wt <- rbind(cbind(wsum(R[n == 27 & method %in% c(CLASSIC, ROBUST)], "heldout_super_acc"), data = "as the article used them"),
            cbind(wsum(W, "heldout_super_acc"), data = "clipped at 3 sd"))
wt[, col := paste0(ifelse(method == CLASSIC, "Classical SOM", "Robust SOM"), ", features ", data)]
wt <- dcast(wt, set ~ col, value.var = "v")
wt <- rbind(wt[match(names(SETS), set)], wt[, c(list(set = "mean"), lapply(.SD, mean)), .SDcols = !"set"])
wt[, set := ifelse(set == "mean", "**Mean of the five**", SETS[set])]
setnames(wt, "set", "Feature set")
w_mean <- function(m) W[method == m & !is.na(heldout_super_acc), .(v = mean(heldout_super_acc)), by = set][, mean(v)]
knitr::kable(wt, digits = 3, caption = sprintf("Accuracy on held-out children with the features as the article used them and clipped at 3 sd (%d seeds)", W[method == CLASSIC & !is.na(heldout_super_acc), uniqueN(seed)]))
```

Clipping lifts the classical SOM from `r f3(kh$heldout_super_acc)` to `r f3(w_mean(CLASSIC))` on held-out children
(mean of the five feature sets), while the robust SOM moves from `r f3(rb$heldout_super_acc)` to `r f3(w_mean(ROBUST))`.
Most of the gain is robustness to outliers, which the robust SOM gets without touching the data or choosing a
threshold`r if (rb$heldout_super_acc > w_mean(CLASSIC)) ", and it stays the more accurate map" else ""`.

# 6. The number of features: Table 5 and beyond

The article compared 24 to 27 features and chose 27. Its saved maps cover 2 to 35 features, one seed each.

```{r curve, echo = FALSE}
C <- res$curve[method %in% c(CLASSIC, ROBUST, "euclid:novel"), .(mean = mean(accuracy), min = min(accuracy), max = max(accuracy)), by = .(set, n, method)]
C[, who := ifelse(method == CLASSIC, "classic", ifelse(method == ROBUST, "robust", "own"))]
C[, set_label := factor(SETS[set], levels = SETS)]
nseeds_curve <- res$curve[method == CLASSIC, uniqueN(seed)]
p_curve(as.data.frame(C), as.data.frame(pubs), c(classic = COL_CLASSIC, robust = COL_ROBUST, own = COL_OWN),
        c(classic = sprintf("Classical SOM, %d seeds", nseeds_curve), robust = sprintf("Robust SOM, %d seeds", nseeds_curve),
          own = sprintf("Online SOM on novel_SOM's own schedule, %d seeds", nseeds_curve)), first = SETS[["tsfresh"]])
rng <- res$curve[method == CLASSIC, .(lo = min(accuracy), hi = max(accuracy)), by = .(set, n)]
inside <- merge(rng, pubs[, .(set, n, accuracy)], by = c("set", "n"))[, mean(accuracy >= lo & accuracy <= hi)]
ahead <- merge(res$curve[method == ROBUST, .(r = mean(accuracy)), by = .(set, n)], res$curve[method == CLASSIC, .(k = mean(accuracy)), by = .(set, n)],
               by = c("set", "n"))[n >= 10]
```

<p class="note">Lines are means over the seeds, bands the range of the seeds; the black line is the article's
published maps. The buttons switch the feature set.</p>

`r sprintf("%.0f%%", 100 * inside)` of the article's published accuracies lie inside the range of
`r nseeds_curve` seeds of the same classical SOM`r if (inside >= 0.6) ", so the jumps of the published curve from one number of features to the next are mostly the luck of one seed" else ""`.
The robust SOM is ahead of the classical one at
`r sum(ahead$r > ahead$k)` of `r nrow(ahead)` combinations of feature set and 10 to 35 features.

```{r table5, echo = FALSE}
t5 <- res$curve[n %in% 24:27 & method %in% c(CLASSIC, ROBUST), .(v = sprintf("%.3f &plusmn; %.3f", mean(accuracy), sd(accuracy))), by = .(set, n, method)]
if (nrow(t5)) {
  t5 <- dcast(t5, set + n ~ method, value.var = "v")
  t5 <- merge(t5, pubs[n %in% 24:27, .(set, n, published = sprintf("%.3f", accuracy))], by = c("set", "n"))
  t5 <- t5[order(match(set, names(SETS)), n), .(`Feature set` = SETS[set], Features = n, `The article (one seed)` = published,
                                                  `Classical SOM` = get(CLASSIC), `Robust SOM` = get(ROBUST))]
  knitr::kable(t5, caption = sprintf("The article's Table 5: accuracy on all data, 24 to 27 features (mean &plusmn; sd over %d seeds)", nseeds_curve))
}
```

# 7. The maps

The article's Fig. 3, redrawn: each unit a pie of the tasks it wins (area by the number of segments), thick lines
between the 7 super-clusters. The classical and the robust SOM are shown at their median seed, a representative
run rather than the best one. Hover a unit for its segments.

```{r map-fits, echo = FALSE}
median_seed <- function(st, m) { x <- R[set == st & n == 27 & method == m]; x$seed[which.min(abs(x$accuracy - median(x$accuracy)))] }
build_maps <- function(st) {
  X <- D$X[[st]][["27"]]; p <- D$models[[27]]$som[[st]]$SOM_results$som_model
  mk <- median_seed(st, CLASSIC); mb <- median_seed(st, ROBUST)
  out <- list(list(title = "The article, published (seed 1245)", fit = list(M = p$codes[[1]], bmu = p$unit.classif, euclid_family = TRUE), pos = p$grid$pts),
              list(title = sprintf("Classical SOM, median seed (%d)", mk), fit = fit_kohonen(X, mk), pos = kohonen::somgrid(5, 5, "hexagonal")$pts),
              list(title = sprintf("Robust SOM, median seed (%d)", mb), fit = fit_som(X, ROBUST, mb), pos = make_grid(5, 5)$pos))
  lapply(out, function(m) { m$super <- super_clusters(m$fit); m$bmu <- m$fit$bmu; m$acc <- article_accuracy(m$bmu, m$super, D$y7)
    m$subtitle <- sprintf("accuracy %.3f", m$acc$accuracy); m$square <- FALSE; m })
}
maps_ts <- build_maps("tsfresh"); maps_sp <- build_maps("spectr")
```

## Spectral and tsfresh features {.tabset .tabset-pills}

### Spectral features

```{r maps-spectral, echo = FALSE}
p_maps(maps_sp, D$y7)
```

### tsfresh features (the article's figure)

```{r maps-tsfresh, echo = FALSE}
p_maps(maps_ts, D$y7)
```

<details><summary>The article's Table 6: segments per super-cluster and task, tsfresh maps</summary>

```{r table6, echo = FALSE, results = "asis"}
for (m in maps_ts) {
  tab <- unclass(m$acc$table); colnames(tab) <- TASKS7
  d6 <- data.frame(`Super-cluster` = seq_len(nrow(tab)), `Named after` = TASKS7[m$acc$task], tab, Total = rowSums(tab),
                   Purity = sprintf("%.2f", apply(tab, 1, max) / pmax(rowSums(tab), 1)), check.names = FALSE)
  cat("\n\n**", m$title, "**, accuracy ", f3(m$acc$accuracy), "\n\n", sep = ""); print(knitr::kable(d6, row.names = FALSE))
}
```

</details>

# 8. Conclusions

<div class="callout">

* **The robust SOM recognises children's movement tasks better than the classical SOM**, with the article's own
  features, map size, clustering and measure. On children it has never seen, the mean accuracy over the five
  feature sets rises from `r f3(kh$heldout_super_acc)` to `r f3(rb$heldout_super_acc)`, and the gain holds on fresh seeds
  and child by child.
* **The gain is largest where the classical map struggles**: spectral `r pts(per_set("heldout_super_acc", ROBUST)[["spectr"]] - per_set("heldout_super_acc", CLASSIC)[["spectr"]])`,
  autocorrelation `r pts(per_set("heldout_super_acc", ROBUST)[["autocorr"]] - per_set("heldout_super_acc", CLASSIC)[["autocorr"]])`, Mix
  `r pts(per_set("heldout_super_acc", ROBUST)[["mix"]] - per_set("heldout_super_acc", CLASSIC)[["mix"]])` points.`r if (spec_beats_all) paste0(" With the robust SOM the spectral features (", f3(ho_rb[["spectr"]]), ") do better than the classical SOM on any feature set, including the article's tsfresh choice (", f3(ho_cl[["tsfresh"]]), ").") else ""`
* **Why:** the features are heavy-tailed, and an L1 winner with a clipped pull keeps outlying segments from
  dragging prototypes. Clipping the features helps the classical SOM too (`r f3(w_mean(CLASSIC))` on held-out
  children), `r if (w_mean(CLASSIC) < rb$heldout_super_acc) "but not as far as the robust SOM goes without clipping" else "about as far as the robust SOM goes"` (`r f3(rb$heldout_super_acc)`).
* **Results from one seed are fragile.** The article's published `r f3(pubs[n == 27 & set == "tsfresh"]$accuracy)` (tsfresh) sits near
  the top of its own spread: `r n_all` seeds of the same call give `r f3(S[n == 27 & set == "tsfresh" & method == CLASSIC & measure == "accuracy"]$mean)` on average, from
  `r f3(S[n == 27 & set == "tsfresh" & method == CLASSIC & measure == "accuracy"]$min)` to
  `r f3(S[n == 27 & set == "tsfresh" & method == CLASSIC & measure == "accuracy"]$max)`. Reporting the mean and spread over seeds,
  on held-out children, makes a SOM study comparable.

</div>

**Limitations.**

* The article chose the 27 features with XGBoost on all 81 children, so the held-out accuracies are optimistic
  for every map. This affects both maps equally and leaves the comparison fair; a fully nested study would choose
  the features inside every fold.
* One data set, 81 children from one study. The child-level test generalises to children like these, not to
  other sensors, ages or tasks.
* On the article's best feature set (tsfresh) the gain is small (`r pts(d_ho[set == "tsfresh"]$diff)` points on
  held-out children), and on RQA the robust SOM is level with the classical one. There the correlation distance
  helps instead (all data `r f3(per_set("accuracy", "correlation:article")[["rqa"]])` against
  `r f3(per_set("accuracy", CLASSIC)[["rqa"]])`), which suggests the scale of the RQA features is the problem.
* The screening tried 34 maps; the confirmation on fresh seeds guards against a lucky choice, but not against a
  data set that happens to favour robust maps.

# 9. Reproduce

```{.bash .fold-show}
Rscript run_som_variants.R        # all maps, about 1.5 hours on 18 cores; results go to results/
```

Then knit `som_variants.Rmd` (a few minutes; `quick: true` in the YAML header knits a small version). The
article's saved models (95 MB) are downloaded to `data/` from its repository on first use and checked by md5.

| File | What it holds |
|---|---|
| `R/article_metrics.R` | the article's accuracy, mnLogLoss and quality measures, ported without caret |
| `R/som_fits.R` | the article's `kohonen::som` call and the novel_SOM variants on both schedules |
| `R/protocols.R` | all data, the article's k-fold, held-out children, per-segment predictions |
| `R/bench.R` | the article's data, the plan, the parallel runs and the cache |
| `R/plots.R`, `R/plots_interactive.R` | the static figures (README) and the interactive ones (this page) |
| `novel_som/` | `ns_core.cpp` and `ns_maps.R` from novel_SOM (commit 2ae89c5), plus the bubble neighbourhood |
| `results/` | the results of the latest run, so the report knits without retraining |

Needs R 4.3 or later with Rcpp and a C++17 compiler (Rtools on Windows), and kohonen, aweSOM, cluster,
data.table, ggplot2, plotly, DT, digest and patchwork.

<details><summary>The analysis code of this page</summary>

```{r ref.label = c("setup", "load", "analysis"), eval = FALSE, echo = TRUE}
```

</details>

<details><summary>Session information</summary>

```{r session, echo = FALSE}
sessionInfo()
```

</details>

```{r save-figures, include = FALSE, eval = params$save_figures && !params$quick}
dir.create("figures", showWarnings = FALSE)
ggsave("figures/som-variants-headline.png", dumbbell_plot(S2[protocol == "heldout_super_acc"],
       title = "Accuracy on children the map has never seen, 27 features",
       subtitle = sprintf("mean over %d seeds; the number is the gain of the robust SOM in percentage points", max(S2[protocol == "heldout_super_acc"]$seeds))),
       width = 9, height = 4.6, dpi = 150, bg = "white")
ggsave("figures/som-variants-children.png", children_plot(PC[set %in% c("spectr", "tsfresh")],
       title = "Child by child: robust minus classical SOM, held-out accuracy",
       subtitle = sprintf("each bar one of the %d children, mean over %d fresh seeds", n_children, n_conf)), width = 10, height = 4.2, dpi = 150, bg = "white")
ggsave("figures/som-variants-curve.png", curve_plot(C[method %in% c(CLASSIC, ROBUST)], pubs,
       setNames(c(COL[["kohonen"]], COL[["article"]]), c(CLASSIC, ROBUST)),
       setNames(c(sprintf("Classical SOM (kohonen::som), %d seeds", nseeds_curve), sprintf("Robust SOM, %d seeds", nseeds_curve)), c(CLASSIC, ROBUST)),
       title = "Accuracy against the number of features", subtitle = "lines: mean over the seeds; bands: range of the seeds; black: the article's published maps, one seed each"),
       width = 11, height = 7, dpi = 150, bg = "white")
spm <- lapply(maps_sp, function(m) hex_pie_map(m$pos, m$bmu, D$y7, m$super, title = m$title, subtitle = paste("Spectral features:", m$subtitle)))
ggsave("figures/som-variants-maps-spectral.png", patchwork::wrap_plots(spm, nrow = 1, guides = "collect") & theme(legend.position = "bottom"),
       width = 15, height = 6.5, dpi = 120, bg = "white")
sm <- dcast(DD[measure %in% c("heldout_super_acc", "accuracy")], set_label ~ measure, value.var = c("classic", "robust", "diff"))
sm_row <- function(lab, x) sprintf("| %s | %.3f | %.3f | **%+.1f** | %.3f | %.3f | %+.1f |", lab, x$classic_heldout_super_acc, x$robust_heldout_super_acc,
                                   100 * x$diff_heldout_super_acc, x$classic_accuracy, x$robust_accuracy, 100 * x$diff_accuracy)
sm_mean <- sm[, lapply(.SD, mean), .SDcols = !"set_label"]
writeLines(c("| Feature set | Held-out children: classical SOM | robust SOM | gain (points) | All 648 segments: classical SOM | robust SOM | gain (points) |",
             "|:--|--:|--:|--:|--:|--:|--:|", sapply(seq_len(nrow(sm)), function(i) sm_row(as.character(sm$set_label[i]), sm[i])),
             sm_row("**Mean of the five**", sm_mean)), "figures/som_variants_summary.md")
lt_md <- copy(lt)[, (4:ncol(lt)) := lapply(.SD, function(v) sprintf("%.3f", v)), .SDcols = 4:ncol(lt)]
writeLines(c("# SOM variants on the article's features", "",
             sprintf("%s: the article's feature matrices (27 features, top XGBoost gain), 5 x 5 maps, 7 PAM super-clusters; all data with %d seeds, held-out children (9 folds of 9 children) with %d seeds. Means over the five feature sets, then held-out accuracy per feature set.",
                     format(Sys.Date()), n_all, n_cv), "", knitr::kable(lt_md, format = "pipe")), "figures/som_variants_table.md")
```
