---
title: "SuperSOM: layered, growing and DTW maps"
subtitle: "Does a layered SOM recognise children's movement tasks better than the best single map? The data and pipeline of Patrikei et al. (2026), *Acta Gymnica*: 28 maps, 10 seeds, children held out"
author: "Andrii Patrikei"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  html_document:
    toc: true
    toc_float:
      collapsed: false
    toc_depth: 2
    code_folding: hide
    theme: flatly
    highlight: tango
params:
  quick: false     # TRUE = 2 seeds and 2 passes, to test the document in minutes (results in supersom/results_quick/)
  cores: 0         # 0 = all but two; used only when supersom/results/ lacks a part of the plan
---

```{css, echo = FALSE}
.main-container { max-width: 1120px; }
body { font-size: 16px; line-height: 1.62; color: #1d1d1b; }
h1.title { font-size: 34px; margin-top: 18px; }
h3.subtitle { font-size: 19px; color: #52514e; line-height: 1.4; }
h1 { font-size: 26px; margin-top: 44px; } h2 { font-size: 20px; margin-top: 30px; }
.lead { font-size: 18px; line-height: 1.6; }
.tiles { display: grid; grid-template-columns: repeat(auto-fit, minmax(210px, 1fr)); gap: 14px; margin: 22px 0 10px; }
.tile { border: 1px solid #e6e5e0; border-left: 4px solid #2a78d6; border-radius: 6px; padding: 12px 16px; background: #fcfcfb; }
.tile.yes { border-left-color: #1baf7a; }
.tile.no { border-left-color: #eb6834; }
.tile .num { font-size: 30px; font-weight: 700; line-height: 1.15; color: #0b0b0b; }
.tile .lbl { font-size: 14px; color: #1d1d1b; margin-top: 4px; }
.tile .sub { font-size: 13px; color: #52514e; margin-top: 2px; }
.callout { border-left: 4px solid #2a78d6; background: #f3f7fd; padding: 12px 18px; margin: 18px 0 22px; border-radius: 0 6px 6px 0; }
.callout p:last-child, .callout ul:last-child { margin-bottom: 0; }
.note { color: #52514e; font-size: 14px; }
table { font-size: 14px; }
.table > thead > tr > th, .table > tbody > tr > td { padding: 5px 9px; }
details summary { cursor: pointer; color: #2a78d6; margin: 8px 0; }
pre code { font-size: 13px; }
span.math.display { display: block; overflow-x: auto; overflow-y: hidden; }
.diagram { margin: 10px 0 18px; }
.diagram svg { width: 100%; max-width: 980px; height: auto; display: block; margin: 0 auto; }
.diagram text { font-family: Lato, "Helvetica Neue", Helvetica, Arial, sans-serif; }
```

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, comment = "#>", fig.align = "center", dpi = 110)
library(data.table)
for (f in c("R/article_metrics.R", "R/som_fits.R", "R/protocols.R", "R/plots.R", "R/plots_interactive.R",
            file.path("supersom", "R", c("ss_maps.R", "ss_protocols.R", "ss_bench.R", "ss_raw.R", "ss_report.R")))) source(f)
load_ss_core()
cores <- if (params$cores > 0) params$cores else max(1L, parallel::detectCores() - 2L)
quick <- isTRUE(params$quick); res_dir <- file.path("supersom", if (quick) "results_quick" else "results")
```

```{r load, include = FALSE}
prepare_ss_cache()                                        # the article's matrices and the raw curves (built once)
S <- ss_data(rlen = if (quick) 2L else 1000L)
plan <- if (quick) ss_plan(1:2, 1:2) else ss_plan()
res <- ss_compute(S, plan, cores, res_dir)
for (p in names(res)) if (!is.null(attr(res[[p]], "failed"))) stop("failed jobs in ", p)
stopifnot(nrow(res$all) == nrow(plan$all), nrow(res$heldout) == 648 * nrow(plan$heldout))
WIN <- pick_winners(res$heldout)
CF <- ss_compute(S, ss_plan_confirm(res, if (quick) 101:102 else 101:120), cores, res_dir)$confirm
LNN_all <- ss_cached("layer_nn", ss_key("layer 1-NN", S$md5, "with predictions"), function() layer_nn(S), res_dir)
LNN <- LNN_all$tab; wrong <- sapply(LNN_all$pred, function(p) p != S$y7)
```

```{r analysis, include = FALSE}
mt <- method_table()
lab <- setNames(mt$label, mt$method); grp <- setNames(mt$group, mt$method)
H <- with_votes(res$heldout)
Sd <- seed_scores(H)
RULES <- c("super", "unit", "vote")
RULE_LAB <- c(super = "held-out children, the article's rule (7 super-clusters)", unit = "held-out children, each unit names its task",
              vote = "held-out children, the layers vote")
SM <- summarise_seeds(Sd, c(RULES, "super_bacc", "unit_bacc", "acc_only_unit", "acc_only_super"))
num <- setdiff(names(res$all)[vapply(res$all, is.numeric, TRUE)], "seed")
AS <- res$all[, lapply(.SD, mean), by = method, .SDcols = num]
mn <- function(m, me) SM[method == m & measure == me]$mean
al <- function(m, me) AS[method == m][[me]]
f3 <- function(x) sprintf("%.3f", x); pts <- function(x) sprintf("%+.1f", 100 * x); pt1 <- function(x) sprintf("%.1f", 100 * abs(x))
n_seeds <- uniqueN(res$heldout$seed); n_all <- uniqueN(res$all$seed); n_conf <- uniqueN(CF$seed); n_children <- uniqueN(S$child)

# the rule fixed in advance: the best unsupervised layered map and the best one-layer map, by the mean held-out
# accuracy under each rule (screening seeds); then both again on fresh seeds
L_S <- WIN$layered_super; L_U <- WIN$layered_unit; O_S <- WIN$single_super; O_U <- WIN$single_unit

# B minus A, seed by seed (the same seeds give the same folds of children): mean, paired t interval, seeds better
paired <- function(a, b, me, D = Sd) {
  x <- merge(D[method == a, .(seed, va = get(me))], D[method == b, .(seed, vb = get(me))], by = "seed"); d <- x$vb - x$va
  ci <- if (nrow(x) > 2 && sd(d) > 0) t.test(d)$conf.int else c(mean(d), mean(d))
  list(diff = mean(d), lo = ci[1], hi = ci[2], better = sum(d > 0), worse = sum(d < 0), seeds = nrow(x))
}
ps <- paired(O_S, L_S, "super"); pu <- paired(O_U, L_U, "unit")
p_glue <- paired("ss_classic_F4", "glue_classic_F4", "super"); p_agree <- paired("ss_rob_F4", "ss_rob_F4_agree", "super")
p_rob <- paired("ss_classic_F4", "ss_rob_F4", "super"); p_sel <- paired("ss_sel7", "ss_sel7_euc", "super")
p_dtw1 <- paired("euc_posture", "dtw_posture", "unit"); p_dtw3 <- paired("ss_R3_euc", "ss_R3_dtw", "unit")
p_sep <- paired("rob_spectr", "vote_sep7", "unit")

# the confirmation on fresh seeds: per child, the mean held-out accuracy over the seeds
CFs <- CF[, .(super = mean(pred_super == y7), unit = mean(pred_unit == y7)), by = .(method, child)]
child_cmp <- function(a, b, rule) {
  x <- merge(CFs[method == a, .(child, a = get(rule))], CFs[method == b, .(child, b = get(rule))], by = "child"); d <- x$b - x$a
  set.seed(1); boot <- replicate(10000, mean(sample(d, replace = TRUE)))
  p <- if (any(d != 0)) suppressWarnings(wilcox.test(x$b, x$a, paired = TRUE, exact = FALSE))$p.value else 1
  list(x = x[, rule := rule], better = sum(d > 0), worse = sum(d < 0), equal = sum(d == 0), gain = mean(d),
       lo = unname(quantile(boot, 0.025)), hi = unname(quantile(boot, 0.975)), p = p)
}
cs <- child_cmp(O_S, L_S, "super"); cu <- child_cmp(O_U, L_U, "unit"); ck <- child_cmp("kohonen_tsfresh", L_S, "super")
conf <- CF[, .(super = mean(pred_super == y7), unit = mean(pred_unit == y7)), by = .(method, seed)][, .(super = mean(super), unit = mean(unit)), by = method]
cfm <- function(m, r) conf[method == m][[r]]

# a sensor goes missing: the 9 children without a gyroscope (all-data maps), and the held-out children mapped with the
# accelerometer curves only; "all7" = the three maps trained on all seven layers, "acc" = the best map trained on
# accelerometer curves alone (by its own held-out accuracy)
ALL7 <- c("ss_all7", "ss_all7_agree", "gss_all7")
ACC_ONLY <- c("dtw_posture", "euc_posture", "dtw_accspec", "ss_A2_dtw")
NG <- res$all[!is.na(nogyro9_unit), .(method, seed, nogyro9 = nogyro9_unit)]
best_acc <- ACC_ONLY[which.max(sapply(ACC_ONLY, mn, me = "unit"))]
ng_mean <- NG[, .(v = mean(nogyro9)), by = method]; ngm <- function(m) ng_mean[method == m]$v
p_ng <- lapply(setNames(ALL7, ALL7), function(m) paired(best_acc, m, "nogyro9", NG))
accH <- H[method %in% c(ALL7, best_acc), .(v = if (method[1] %in% ALL7) mean(!is.na(acc_unit) & acc_unit == y7) else mean(pred_unit == y7)), by = .(method, child)]
acc_child <- function(m) { x <- merge(accH[method == best_acc, .(child, a = v)], accH[method == m, .(child, b = v)], by = "child"); d <- x$b - x$a
  list(better = sum(d > 0), worse = sum(d < 0), p = suppressWarnings(wilcox.test(x$b, x$a, paired = TRUE, exact = FALSE))$p.value) }
ac7 <- acc_child("ss_all7")

# growing against fixed, the same layers and seeds (all data)
GF <- rbindlist(lapply(list(c("ss_rob_F4", "gss_rob_F4"), c("ss_R3_dtw", "gss_R3_dtw"), c("ss_all7", "gss_all7"), c("ss_sel7", "gss_sel7")), function(p) {
  x <- merge(res$all[method == p[1]], res$all[method == p[2]], by = "seed", suffixes = c("_f", "_g"))
  hs <- paired(p[1], p[2], "super"); hu <- paired(p[1], p[2], "unit")
  data.table(fixed = p[1], growing = p[2], te_f = mean(x$TE_own_f), te_g = mean(x$TE_own_g), te_lower = sum(x$TE_own_g < x$TE_own_f),
             tp_f = mean(x$TP_f), tp_g = mean(x$TP_g), qe_f = mean(x$QE_own_f), qe_g = mean(x$QE_own_g), sec_f = mean(x$seconds_f), sec_g = mean(x$seconds_g),
             ho_super = hs$diff, ho_super_better = hs$better, ho_unit = hu$diff, seeds = nrow(x))
}))
sv <- range(1 - GF$sec_g / GF$sec_f); speed <- sprintf("%.0f to %.0f%%", 100 * sv[1], 100 * sv[2])

# fresh seeds, seed by seed (the paired test of the screening, on seeds 101-120)
CFseed <- CF[, .(super = mean(pred_super == y7), unit = mean(pred_unit == y7)), by = .(method, seed)]
pcu <- paired(O_U, L_U, "unit", CFseed); pcs <- paired(O_S, L_S, "super", CFseed)
# which layers the folds of the pre-set winner kept: "spectr x4; spectr+tsfresh x2; ..." per seed
fold_sets <- function(D, m) {
  x <- unique(D[method == m, .(seed, layers)])
  rbindlist(lapply(x$layers, function(l) rbindlist(lapply(strsplit(l, "; ")[[1]], function(p) {
    s2 <- strsplit(p, " x")[[1]]; data.table(set = s2[1], n = as.integer(s2[2])) }))))[, .(n = sum(n)), by = set]
}
fs_h <- fold_sets(res$heldout, L_S); fs_c <- fold_sets(CF, L_S)
alone_h <- fs_h[set == "spectr", sum(n)]; alone_c <- fs_c[set == "spectr", sum(n)]
nlay <- function(fs) fs[, .(k = lengths(strsplit(set, "+", fixed = TRUE)), n)][, .(n = sum(n)), by = k]
nl_dtw <- nlay(fold_sets(res$heldout, "ss_sel7")); dtw_12 <- nl_dtw[k %in% 2:3, sum(n)]
# DTW against sum of squares, the same layers
p_dtw1s <- paired("euc_posture", "dtw_posture", "super"); p_dtw_sel <- paired("ss_sel7_euc", "ss_sel7", "super")
p_adapt <- paired("ss_rob_F4", "ss_rob_F4_adapt", "super")
# the agreement weights: RQA's weight at the end against kohonen's start weight (the same start rows, seed by seed)
wnum <- function(m) { x <- res$all[method == m, .(seed, w)]; x[, rqa := as.numeric(sapply(strsplit(w, " "), `[`, 1))]; x[, .(seed, rqa)] }
wr <- merge(wnum("ss_rob_F4"), wnum("ss_rob_F4_agree"), by = "seed", suffixes = c("_0", "_1"))[, rel := rqa_1 / rqa_0 - 1]
# the layer vote against the unit rule, layered maps
vu <- dcast(SM[measure %in% c("vote", "unit") & method %in% mt[!group %in% c("single", "separate")]$method], method ~ measure, value.var = "mean")[, d := vote - unit]
# the floating radius under the bubble neighbourhood: the same growing map with it switched off
g4 <- SS_METHODS[["gss_rob_F4"]]
float_diff <- max(sapply(1:2, function(sd) max(abs(fit_ss(S$XL, g4, sd, rlen = S$rlen)$M - fit_ss(S$XL, modifyList(g4, list(float = 0)), sd, rlen = S$rlen)$M))))
```

<p class="lead">
A SuperSOM (<code>kohonen::supersom</code>) trains <b>one</b> map on several blocks of data at once, the "layers": every
layer keeps its own distance and weight, and the winning unit is the one closest over all of them. Here every version
of it was built from the best ingredients of the earlier reports (the robust Manhattan SOM, the DTW-SOM, and my
earlier growing DTW-SOM) and tested the way <code>som_variants.Rmd</code> tested the single maps: the article's 648 segments,
its 5&times;5 map and 7 super-clusters, children the map has never seen, `r n_seeds` seeds, and a confirmation on
`r n_conf` fresh seeds.
</p>

<div class="tiles">
<div class="tile no"><div class="num">`r pts(ps$diff)` points</div>
<div class="lbl">best layered map against the best single map, held-out children, the article's rule</div>
<div class="sub">`r f3(mn(L_S, "super"))` against `r f3(mn(O_S, "super"))`; worse on `r ps$worse` of `r ps$seeds` seeds; fresh seeds: `r cs$worse` children worse, `r cs$better` better</div></div>
<div class="tile no"><div class="num">`r pts(pu$diff)` points</div>
<div class="lbl">the same, when each unit names its task</div>
<div class="sub">`r f3(mn(L_U, "unit"))` against `r f3(mn(O_U, "unit"))`; small but real: lower on `r pcu$worse` of `r pcu$seeds` fresh seeds (interval `r pts(pcu$lo)` to `r pts(pcu$hi)`)</div></div>
<div class="tile yes"><div class="num">`r pts(ngm("ss_all7") - ngm(best_acc))` points</div>
<div class="lbl">the 9 children the article dropped (gyroscope failed), recognised from the accelerometer alone</div>
<div class="sub">`r f3(ngm("ss_all7"))` with a map trained on all 7 layers, `r f3(ngm(best_acc))` with the best accelerometer-only map; better on `r p_ng$ss_all7$better` of `r p_ng$ss_all7$seeds` seeds</div></div>
<div class="tile"><div class="num">partly new</div>
<div class="lbl">a layered map whose lattice grows during training</div>
<div class="sub">kernel form published (a growing multiple-kernel GSOM, Wijewardena et al. 2015); kohonen's form with DTW and Manhattan layers and medoid births: not found</div></div>
</div>

<div class="callout">

**Does it work?**

* **For the article's question, no.** With every sensor available, the best single feature set on the robust SOM
  (`r lab[[O_S]]`, `r f3(mn(O_S, "super"))`; `r lab[[O_U]]`, `r f3(mn(O_U, "unit"))` when each unit names its task) stays
  ahead of every layered map. Layering the article's four feature sets costs
  `r pt1(mn("ss_rob_F4", "super") - mn("rob_spectr", "super"))` points against the spectral map alone; letting each fold choose its own layers wins most of it
  back (`r f3(mn(L_S, "super"))`) but does not overtake.
* **Where it works: a missing sensor.** A map trained on all seven layers and fed only the accelerometer curves recognises
  `r f3(ngm("ss_all7"))` of the segments of the 9 children whose gyroscope failed, against `r f3(ngm(best_acc))` for the best map
  trained on the accelerometer curves alone. Only a layered map can do this: train with everything, map with what you have.
* **Growing adds nothing to the accuracy, and DTW costs some in layered maps.** The growing map reaches the same accuracy,
  is `r speed` faster and less well ordered. DTW lifts a single posture-curve map (`r pts(p_dtw1$diff)` points when each unit
  names its task, `r pts(p_dtw1s$diff)` under the article's rule) but lowers the layered maps: `r pts(p_dtw3$diff)` points for the
  three raw curves (`r p_dtw3$worse` of `r p_dtw3$seeds` seeds), `r pts(p_dtw_sel$diff)` for the layer-choosing map.

</div>

# 1. What a SuperSOM is

<div class="diagram">
<svg viewBox="0 0 980 300" xmlns="http://www.w3.org/2000/svg" role="img" aria-label="Seven layers, each with its own distance, summed with weights; the unit with the smallest sum wins on one hexagonal map">
<g font-size="13" fill="#1d1d1b">
<text x="10" y="18" font-weight="700" fill="#52514e">one segment, seven layers</text>
<g stroke="#e6e5e0" stroke-width="1.5">
<rect x="10" y="30" width="250" height="28" rx="4" fill="#fdeee8"/><rect x="10" y="62" width="250" height="28" rx="4" fill="#fdeee8"/>
<rect x="10" y="94" width="250" height="28" rx="4" fill="#fdeee8"/><rect x="10" y="126" width="250" height="28" rx="4" fill="#fdeee8"/>
<rect x="10" y="168" width="250" height="28" rx="4" fill="#e6f6ef"/><rect x="10" y="200" width="250" height="28" rx="4" fill="#e6f6ef"/>
<rect x="10" y="232" width="250" height="28" rx="4" fill="#e6f6ef"/></g>
<text x="20" y="49">RQA features (27)</text><text x="20" y="81">Autocorrelation features (27)</text>
<text x="20" y="113">Spectral features (27)</text><text x="20" y="145">tsfresh features (27)</text>
<text x="20" y="187">Posture curve (3 axes x 64)</text><text x="20" y="219">Accelerometer spectrum (3 x 40)</text>
<text x="20" y="251">Gyroscope spectrum (3 x 40)</text>
<text x="270" y="95" fill="#52514e">Manhattan, clipped pull</text><text x="270" y="219" fill="#52514e">DTW, warped pull</text>
<g stroke="#52514e" stroke-width="1.2" fill="none"><path d="M262 44 L430 148"/><path d="M262 76 L430 148"/><path d="M262 108 L430 148"/><path d="M262 140 L430 148"/>
<path d="M262 182 L430 148"/><path d="M262 214 L430 148"/><path d="M262 246 L430 148"/></g>
<circle cx="470" cy="148" r="40" fill="#fcfcfb" stroke="#2a78d6" stroke-width="2"/>
<text x="470" y="144" text-anchor="middle" font-size="15">&#931; w&#8343; d&#8343;</text><text x="470" y="163" text-anchor="middle" font-size="11" fill="#52514e">weighted sum</text>
<path d="M512 148 L600 148" stroke="#52514e" stroke-width="1.5" marker-end="url(#ar)"/>
<defs><marker id="ar" markerWidth="8" markerHeight="8" refX="7" refY="4" orient="auto"><path d="M0,0 L8,4 L0,8 z" fill="#52514e"/></marker></defs>
<text x="520" y="138" font-size="11" fill="#52514e">smallest wins</text>
<g stroke="#ffffff" stroke-width="2" fill="#dfe9f7">
<polygon points="640,96 657,86 674,96 674,116 657,126 640,116"/><polygon points="674,96 691,86 708,96 708,116 691,126 674,116"/>
<polygon points="708,96 725,86 742,96 742,116 725,126 708,116"/><polygon points="657,126 674,116 691,126 691,146 674,156 657,146"/>
<polygon points="691,126 708,116 725,126 725,146 708,156 691,146" fill="#2a78d6"/><polygon points="725,126 742,116 759,126 759,146 742,156 725,146"/>
<polygon points="640,156 657,146 674,156 674,176 657,186 640,176"/><polygon points="674,156 691,146 708,156 708,176 691,186 674,176"/>
<polygon points="708,156 725,146 742,156 742,176 725,186 708,176"/><polygon points="657,186 674,176 691,186 691,206 674,216 657,206"/>
<polygon points="691,186 708,176 725,186 725,206 708,216 691,206"/><polygon points="725,186 742,176 759,186 759,206 742,216 725,206"/></g>
<text x="640" y="240" fill="#52514e">one map: every unit has a prototype</text><text x="640" y="256" fill="#52514e">in every layer</text>
<text x="790" y="110" font-size="12" fill="#52514e">the winner and its</text><text x="790" y="126" font-size="12" fill="#52514e">neighbours move toward</text>
<text x="790" y="142" font-size="12" fill="#52514e">the segment, every layer</text><text x="790" y="158" font-size="12" fill="#52514e">by its own pull</text>
</g></svg>
</div>

$$c(\mathbf{x}) = \arg\min_k \sum_{l} w_l\, d_l(\mathbf{x}_l, \mathbf{m}_{k,l}), \qquad \mathbf{m}_{j,l} \leftarrow \mathbf{m}_{j,l} + \alpha(t)\, h_{cj}(t)\, \Delta_l(\mathbf{x}_l, \mathbf{m}_{j,l})$$

Each layer $l$ has its own distance $d_l$, its own weight $w_l$ and its own pull $\Delta_l$. By default, as in kohonen, the
weights make every layer count the same on average: $w_l \propto 1/\bar d_l$, with $\bar d_l$ the mean distance between the
starting units in layer $l$. The new compiled core (`supersom/ss_core.cpp`) is checked against kohonen itself:

```{r identity, echo = FALSE}
chk1 <- sapply(1:3, function(sd) max(abs(fit_kohonen(unclass(S$XL$tsfresh), sd)$M - fit_ss(S$XL, ss_method("k", "", "tsfresh", "classic"), sd, rlen = S$rlen)$M)))
F4l <- S$XL[F4]; set.seed(1); idx <- sample(648, 25, replace = TRUE)
ks <- kohonen::supersom(lapply(F4l, unclass), grid = kohonen::somgrid(5, 5, "hexagonal"), rlen = S$rlen, alpha = c(0.05, 0.01),
                        init = lapply(F4l, function(x) unclass(x)[idx, ]), dist.fcts = "sumofsquares", maxNA.fraction = 8)
chk2 <- max(abs(do.call(cbind, ks$codes) - fit_ss(S$XL, ss_method("s", "", F4, "classic"), 1, rlen = S$rlen)$M))
knitr::kable(data.frame(Check = c("One layer against kohonen::som (the article's call), seeds 1 to 3",
                                  "Four layers against kohonen::supersom (sum of squares, kohonen's layer weights), seed 1"),
                        `Largest difference between the codebooks` = c(max(chk1), chk2), check.names = FALSE))
```

Every map below is therefore an exact `kohonen::supersom` (with the article's random start rows passed as `init`) when
its layers use kohonen's sum of squares, and the same loop with other distances, pulls, weights or a growing lattice
otherwise. One consequence is worth knowing: with sum-of-squares layers a SuperSOM *is* a plain SOM on the layers glued
side by side, each multiplied by $\sqrt{w_l}$ (winner and update alike). The plain glued table, every layer weighted the
same, differs from the SuperSOM only by kohonen's layer weights, and by `r pts(p_glue$diff)` points on held-out children
(`r p_glue$better` seeds better, `r p_glue$worse` worse): layers only matter when they bring their own distance, weight or pull.

# 2. The layers

The article's four feature sets (Mix, its fifth, is the first three together) and three curves built from the raw IMU
signal (`supersom/R/ss_raw.R`), each on an axis where an elastic alignment means something: the trunk's posture over the
course of the task, and the accelerometer and gyroscope spectra, along which DTW lets a child step faster or slower. How
well each layer separates the tasks by itself, before any map (a segment takes the task of its nearest segment of
another child):

```{r layers, echo = FALSE}
LNN_t <- LNN[, .(Layer = LAYER_LABELS[layer], What = ifelse(kind == "features", "the article's 27 features (top XGBoost gain, scaled)",
                                                           c(posture = "accelerometer, low-passed, 64 points over the task",
                                                             accspec = "log power, 0.25 to 10 Hz, 40 bins per axis",
                                                             gyrospec = "log power, 0.25 to 10 Hz, 40 bins per axis")[layer]),
                 Columns = columns, `Distance in the maps` = ifelse(kind == "features", "Manhattan", "DTW"),
                 `1-NN, Euclidean` = euclid, `1-NN, own distance` = own)]
knitr::kable(LNN_t, digits = 3, caption = "Leave-one-child-out 1-NN accuracy of every layer on its own")
```

Every layer carries the tasks, and the layers err on different segments: the spectral and tsfresh features are wrong
on `r sum(wrong[, "spectr"])` and `r sum(wrong[, "tsfresh"])` segments, only `r sum(wrong[, "spectr"] & wrong[, "tsfresh"])` of them
shared, and the posture curve gets `r sum(wrong[, "spectr"] & !wrong[, "posture"])` of the spectral errors right. So a
combination could in principle beat the best layer. The question is whether one map can use that.

# 3. The test

| | Held fixed (the article's) | Changed |
|---|---|---|
| Data | 648 segments, 81 children, 7 tasks; the article's feature matrices | plus three raw curves |
| Map | 5&times;5 units (the growing maps grow to 25), 1000 passes, bubble neighbourhood, learning rate 0.05 to 0.01 | the layers, their distances, weights, growth |
| Clusters | 7 PAM super-clusters, each named after its majority task | PAM on the map's own distance (Euclidean for sum-of-squares maps, as the article) |
| Judged | held-out children: 9 folds of 9 children, `r n_seeds` seeds; all data, `r n_all` seeds | plus unit majority, layer vote, a missing sensor |

The 28 maps were fixed before the full run (after a pilot of three seeds on all data and one held-out seed). They fall
into one-layer controls, layered maps of the four feature sets, of the three raw curves, of all seven layers, and of the
layers each fold chooses for itself: a greedy forward selection by leave-one-child-out 1-NN accuracy, run on the 72
training children of the fold only. The winner among the layered maps and the best one-layer map are picked by a rule
fixed in advance (best mean held-out accuracy under each rule) and run again on `r n_conf` fresh seeds.

<details><summary>The 28 maps</summary>

```{r methods, echo = FALSE}
MT <- mt[, .(Map = label, Group = GROUPS[group], Layers = gsub("\\+", " + ", layers), Grows = ifelse(grow, "yes", ""))]
dt_table(MT, page = 30, filter = "none")
```

</details>

# 4. The verdict

```{r verdict, echo = FALSE}
p_verdict(SM, mt, RULES, RULE_LAB, c("Article's rule (7 super-clusters)", "Each unit names its task", "The layers vote"))
```

<p class="note">Dots: the mean over `r n_seeds` seeds; whiskers: the range of the seeds. The dashed line is the best
one-layer map under the chosen rule; hover a dot for the gain against it. The buttons switch the rule and re-sort the maps.
"The layers vote": every layer of a layered map finds its own nearest unit and the units' tasks vote.</p>

* **No layered map beats the best one-layer map** under any rule. The pre-set winner among the layered maps is
  `r lab[[L_S]]`: `r f3(mn(L_S, "super"))` on held-out children under the article's rule, `r pts(ps$diff)` points against
  `r lab[[O_S]]` (95% interval `r pts(ps$lo)` to `r pts(ps$hi)`, worse on `r ps$worse` of `r ps$seeds` seeds); each unit
  naming its task, `r f3(mn(L_U, "unit"))` against `r f3(mn(O_U, "unit"))` (`r pts(pu$diff)`, interval `r pts(pu$lo)` to `r pts(pu$hi)`).
* **The four feature sets together are worse than the best of them**: `r f3(mn("ss_rob_F4", "super"))` for the robust
  SuperSOM, `r f3(mn("ss_classic_F4", "super"))` for `kohonen::supersom`. The robust layers help
  (`r pts(p_rob$diff)` points), learned agreement weights a little more (`r pts(p_agree$diff)`, `r p_agree$better` of `r p_agree$seeds` seeds),
  and the inverse-error weights tend to make it worse (`r f3(mn("ss_rob_F4_adapt", "super"))`, `r pts(p_adapt$diff)` points,
  interval `r pts(p_adapt$lo)` to `r pts(p_adapt$hi)`).
* **Voting does not rescue it either**: the layer vote lifts a layered map a little above its unit rule (at most
  `r pts(max(vu$d))` points) but never above the spectral map alone, and seven separate one-layer maps voting reach
  `r f3(mn("vote_sep7", "unit"))` against `r f3(mn("rob_spectr", "unit"))`.

<details><summary>Every map and every measure (sortable, searchable, downloadable)</summary>

```{r leaderboard, echo = FALSE}
wide <- dcast(SM[measure %in% c(RULES, "super_bacc", "acc_only_unit")], method ~ measure, value.var = "mean")
lb <- merge(merge(mt, wide, by = "method", all.x = TRUE), AS, by = "method", all.x = TRUE)[order(-super)]
LB <- lb[, .(Map = label, Group = GROUPS[group], `Held-out, super-clusters` = super, `Held-out, units` = unit, `Held-out, layer vote` = vote,
             `Held-out, balanced` = super_bacc, `Held-out, accelerometer only` = acc_only_unit, `All data, super-clusters` = accuracy,
             `All data, units` = unit_acc_in, ARI = ari_super, NMI = nmi_super, Purity = purity, `QE (own)` = QE_own, `TE (own)` = TE_own,
             `Topographic product` = TP, `9 children, no gyroscope` = nogyro9_unit, `Seconds per map` = seconds)]
dt_table(LB, round_cols = names(LB)[3:ncol(LB)], page = 30)
```

<p class="note">All-data measures: one map on all 648 segments, `r n_all` seeds. QE and TE in the map's own weighted
distance (comparable only between maps on the same layers); the topographic product (Bauer &amp; Pawelzik) on the
prototypes against the lattice; ARI and NMI of the 7 super-clusters against the tasks; purity: the share of segments in
the majority task of their unit. Held-out, balanced: every task weighs the same (crawling has twice the segments).</p>

</details>

# 5. Why layering loses

```{r purity, echo = FALSE}
p_purity_vs_super(AS, mt)
```

<p class="note">Every dot one map, the mean over `r n_all` seeds on all 648 segments.</p>

The units of the layered feature-set maps are nearly as pure as those of the best one-layer map (right side of the chart):
when every unit names its task, the four-set SuperSOM gets `r f3(al("ss_rob_F4", "unit_acc_in"))` of the segments right,
the spectral map `r f3(al("rob_spectr", "unit_acc_in"))`. What breaks is the
step the article adds on top, the 7 super-clusters: PAM groups the 25 prototypes by their distances, and in a layered map
those distances are the weighted sum over all layers. A weak layer counts as much as a strong one (that is what the
default weights do), so RQA, the weakest layer (1-NN `r f3(LNN[layer == "rqa"]$own)`; its own map `r f3(mn("rob_rqa", "unit"))` of the held-out
segments when each unit names its task, `r f3(mn("rob_rqa", "super"))` under the article's rule), blurs the
distances between the prototypes of different tasks. Choosing the layers per fold removes the weak ones. The folds always keep the spectral features; with DTW curves they
mostly add one or two more layers (`r dtw_12` of `r 9 * n_seeds` folds), while the pre-set winner (curves by sum of
squares) kept the spectral features *alone* in `r alone_h` of `r 9 * n_seeds` folds, where it simply is the spectral
one-layer map:

```{r chosen, echo = FALSE}
ch <- unique(res$heldout[method %in% c("ss_sel7", "ss_sel7_euc"), .(method, seed, layers)])
cnt <- rbindlist(lapply(seq_len(nrow(ch)), function(i) {
  parts <- strsplit(ch$layers[i], "; ")[[1]]
  rbindlist(lapply(parts, function(p) { s <- strsplit(p, " x")[[1]]; data.table(method = ch$method[i], set = s[1], n = as.integer(s[2])) }))
}))
cnt <- cnt[, .(folds = sum(n)), by = .(method, set)]
cnt[, share := folds / sum(folds), by = method]
layer_share <- rbindlist(lapply(POOL7, function(l) cnt[, .(layer = l, share = sum(share[grepl(paste0("(^|\\+)", l, "(\\+|$)"), set)])), by = method]))
knitr::kable(dcast(layer_share, layer ~ method, value.var = "share")[match(POOL7, layer)][, layer := LAYER_LABELS[layer]],
             digits = 2, col.names = c("Layer", "chosen, curves with DTW", "chosen, curves by sum of squares"),
             caption = sprintf("Share of the %d folds (%d seeds x 9) whose training children chose each layer", 9 * n_seeds, n_seeds))
```

<details><summary>How the learned weights move</summary>

```{r weights, echo = FALSE}
WTr <- ss_cached("weight_trace", ss_key("weight trace", S$md5, S$rlen), function() rbindlist(lapply(c("ss_rob_F4_agree", "ss_rob_F4_adapt"), function(id) {
  f <- fit_ss(S$XL, SS_METHODS[[id]], 1, rlen = S$rlen, trace = TRUE); w <- f$w_trace
  rbindlist(lapply(seq_along(F4), function(l) data.table(method = id, layer = F4[l], pass = w[, "step"] / 648, w = w[, l + 1])))
})), res_dir)
w0 <- as.numeric(strsplit(res$all[method == "ss_rob_F4" & seed == 1]$w, " ")[[1]])  # kohonen's start weights, seed 1
WTr <- rbind(WTr, rbindlist(lapply(c("ss_rob_F4_agree", "ss_rob_F4_adapt"), function(id) data.table(method = id, layer = F4, pass = 0, w = w0))))[order(method, layer, pass)]
p_weights(WTr, c(ss_rob_F4_agree = "agreement", ss_rob_F4_adapt = "inverse error"))
```

<p class="note">Seed 1, all data, the weights after every pass (pass 0: kohonen's start weights). Inverse error ($w_l \propto$ the layer's length over its
mean distance to the winner, the maximum-likelihood noise scale) keeps RQA high: its features are correlated and easy to
fit, and easy to fit is not the same as informative. Agreement ($w_l \propto w_l^0 / (1 + g_l)$, $g_l$ the mean lattice distance, over all segments, between the layer's own
nearest unit and the winner, recomputed after every pass; a per-pass version of a running-mean rule I tried
in novel_SOM, not published) moves weight away from RQA, mostly in the first pass: seed 1 from `r f3(wr$rqa_0[1])`
(kohonen's start weight, pass 0) to `r f3(wr$rqa_1[1])`; `r sprintf("%.0f to %.0f%%", -100 * max(wr$rel), -100 * min(wr$rel))`
less over the `r nrow(wr)` seeds.</p>

</details>

# 6. Fresh seeds, child by child

The rule picked `r lab[[L_S]]` (layered) and `r lab[[O_S]]` / `r lab[[O_U]]` (one layer). On `r n_conf` fresh seeds (new
starts, new orders, new folds of children), each child's held-out accuracy averaged over the seeds:

```{r children, echo = FALSE}
CCd <- rbind(cs$x, cu$x)
p_child_bars(CCd, "one layer", "layered", c("super", "unit"), c(sprintf("Article's rule, against %s", lab[[O_S]]), sprintf("Unit rule, against %s", lab[[O_U]])),
             c("Article's rule", "Each unit names its task"))
```

```{r children-table, echo = FALSE}
ctab <- rbindlist(lapply(list(list(cs, O_S, "the article's rule"), list(cu, O_U, "each unit names its task"), list(ck, "kohonen_tsfresh", "the article's rule")), function(z)
  data.table(Rule = z[[3]], `Against` = lab[[z[[2]]]], `Children better / worse / equal` = sprintf("%d / %d / %d", z[[1]]$better, z[[1]]$worse, z[[1]]$equal),
             `Mean gain per child (points)` = sprintf("%s (%s to %s)", pts(z[[1]]$gain), pts(z[[1]]$lo), pts(z[[1]]$hi)),
             `Wilcoxon p` = ifelse(z[[1]]$p < 1e-4, "< 0.0001", sprintf("%.4f", z[[1]]$p)))))
knitr::kable(ctab, caption = sprintf("%s on %d fresh seeds, child by child (bootstrap interval over the %d children; paired Wilcoxon test)", lab[[L_S]], n_conf, n_children))
```

The fresh seeds repeat the screening: under the article's rule the layered map is `r f3(cfm(L_S, "super"))` against
`r f3(cfm(O_S, "super"))`, worse for `r cs$worse` children and better for `r cs$better`. Each unit naming its task it is
`r f3(cfm(L_U, "unit"))` against `r f3(cfm(O_U, "unit"))`: no difference child by child (p = `r sprintf("%.2f", cu$p)`), but lower
on `r pcu$worse` of `r pcu$seeds` fresh seeds (`r pts(pcu$diff)` points, interval `r pts(pcu$lo)` to `r pts(pcu$hi)`), a small
deficit rather than a tie. Part of the closeness is no contest at all: in `r alone_c` of the `r 9 * n_conf` fresh folds the
training children chose the spectral features alone, and there the layered map *is* the spectral one-layer map. It does beat the article's own map
(`kohonen::som` on tsfresh, `r f3(cfm("kohonen_tsfresh", "super"))`) for `r ck$better` children against `r ck$worse`, but so does
the robust one-layer map of `som_variants.Rmd`.

# 7. A sensor goes missing

The article dropped 9 of its 90 children because their gyroscope recorded nothing. Their accelerometer is complete, so
the posture curve and the accelerometer spectrum exist for them; the article's features (built from all six axes) and the
gyroscope spectrum do not. A layered map can still place such a child: it finds the winner with the layers the child has
(kohonen's `whatmap`) and names it from the units, which were trained on everything.

```{r missing, echo = FALSE}
MS <- merge(merge(SM[measure == "unit", .(method, all_layers = mean)], SM[measure == "acc_only_unit", .(method, acc_only = mean)], by = "method"),
            ng_mean[, .(method, nogyro9 = v)], by = "method")
MS[, label := lab[method]]
p_missing(MS)
```

<p class="note">Each unit names its task. Grey: held-out children with every layer the map has; aqua: the same children
mapped with the accelerometer curves only, as if their gyroscope had failed; magenta: the 72 segments of the 9 real
children without a gyroscope, mapped by maps trained on all 81 children. Means over `r n_seeds` seeds. A map whose fold
chose no accelerometer curve cannot place a child without a gyroscope; those segments count as wrong (the low aqua dot of
the layer-choosing map with curves by sum of squares).</p>

* **The maps trained on all seven layers recognise the 9 dropped children best**: `r f3(ngm("ss_all7"))`,
  `r f3(ngm("ss_all7_agree"))` with agreement weights and `r f3(ngm("gss_all7"))` growing, against `r f3(ngm(best_acc))` for the
  best map trained on accelerometer curves alone (`r lab[[best_acc]]`): `r pts(p_ng$ss_all7$diff)` points, better on
  `r p_ng$ss_all7$better` of `r p_ng$ss_all7$seeds` seeds (interval over the seeds `r pts(p_ng$ss_all7$lo)` to `r pts(p_ng$ss_all7$hi)`;
  it covers the training runs on the same 72 segments, not other children). These 9 children were in no step of the
  article's feature selection, and the maps are trained on the other 81.
* **The simulation on the 81 children agrees**: mapped with the accelerometer curves only, the seven-layer map gets
  `r f3(mn("ss_all7", "acc_only_unit"))` against `r f3(mn(best_acc, "unit"))`, better for `r ac7$better` children and worse for
  `r ac7$worse` (paired Wilcoxon p = `r sprintf("%.3f", ac7$p)`). This comparison leans toward the seven-layer map: its
  feature layers come from the article's XGBoost selection on all 81 children, labels included, which the raw-curve map
  never uses. The 9 real children above are free of that.
* Why: the units are cut by all the evidence, so the accelerometer prototypes of a unit belong to one task more cleanly
  than in a map that only ever saw the accelerometer. The price is `r pt1(mn("ss_all7", "unit") - mn("ss_all7", "acc_only_unit"))` points
  against the same map with every layer (`r f3(mn("ss_all7", "unit"))`).

# 8. The growing map

The growing maps are my earlier growing DTW-SOM (not published) with the SuperSOM's distance: seven units to start, then a birth every
`r round(0.6 * S$rlen / 18)` passes (`r format(round(0.6 * S$rlen * 648 / 18), big.mark = ",")` steps on all 648 segments)
until 60% of the training, the radius held at one step while the map grows; each child is the medoid (a real segment, in
every layer at once) of the half of its parent's segments that the parent fits worst; they end with 25 units, like the
fixed maps. Its floating radius is switched on but does nothing here: with kohonen's bubble neighbourhood on whole
steps of the lattice, a radius between 1 and 2 moves the same units as a radius of 1 (with and without it, the growing
map of the four feature sets ends with `r if (float_diff == 0) "identical codebooks" else sprintf("codebooks that differ by at most %.2g", float_diff)`, seeds 1 and 2).

```{r growth, echo = FALSE}
gm <- SS_METHODS[["gss_all7"]]
G <- ss_cached("growth_trace", ss_key("growth trace", S$md5, S$rlen), function() {
  g <- grow_traced(S$XL, gm, 1, rlen = S$rlen)
  g$same_as_fit <- max(abs(g$final$M - fit_ss(S$XL, gm, 1, rlen = S$rlen)$M))
  g
}, res_dir)
p_growth(G, S$y7, S$rlen * length(S$y7))
```

<p class="note">`r lab[["gss_all7"]]`, seed 1, all 648 segments. Every hexagon a unit, coloured by the task most of its
segments belong to; the star marks the newborn unit. Press Play or drag the slider. The traced run is the same training
as the benchmark's (largest difference between the final codebooks: `r G$same_as_fit`).</p>

```{r grow-table, echo = FALSE}
knitr::kable(GF[, .(Layers = lab[fixed], `TE fixed / growing` = sprintf("%.3f / %.3f", te_f, te_g), `Seeds with lower TE` = sprintf("%d of %d", te_lower, seeds),
                    `Topographic product` = sprintf("%.3f / %.3f", tp_f, tp_g), `Seconds per map` = sprintf("%.1f / %.1f", sec_f, sec_g),
                    `Held-out, super-clusters` = pts(ho_super), `Held-out, units` = pts(ho_unit))],
             caption = "Growing against fixed, the same layers (all data for the map measures; held-out gains in points, growing minus fixed)")
```

Growing does not change the accuracy beyond the seed noise; it saves `r speed` of the time (fewer units to compare while it
grows) and orders the map slightly *worse* on the article's schedule: TE higher on average in all four pairs, and on
most seeds in `r sum(GF$te_lower < GF$seeds / 2)` of them. On a Gaussian
schedule my growing DTW-SOM ordered maps better; with kohonen's bubble neighbourhood, whose radius
drops to the winner alone right after the growth, that advantage does not carry over.

# 9. The maps

```{r map-fits, echo = FALSE}
median_seed <- function(m) { x <- res$all[method == m]; x$seed[which.min(abs(x$accuracy - median(x$accuracy)))] }
MP <- ss_cached("maps", ss_key("maps v2", S$md5, S$rlen), function() lapply(c(O_S, L_S, "gss_all7"), function(m) {
  meth <- with_selection(SS_METHODS[[m]], S$XL, S$child, S$y7); sd <- median_seed(m)
  f <- fit_ss(S$XL, meth, sd, rlen = S$rlen, y = S$y7); sup <- super_clusters_ss(f)
  list(title = sprintf("%s (seed %d)", c(lab[[O_S]], "Layered: chosen layers", "Growing: all 7 layers")[match(m, c(O_S, L_S, "gss_all7"))], sd), pos = f$pos, bmu = f$bmu, super = sup, acc = article_accuracy(f$bmu, sup, S$y7), square = FALSE,
       layers = paste(LAYER_LABELS[meth$layers], collapse = ", "))
}), res_dir)
MP <- lapply(MP, function(m) { m$subtitle <- sprintf("accuracy %.3f (all data)", m$acc$accuracy); m })
p_maps(MP, S$y7)
```

<p class="note">The article's Fig. 2 for three maps at their median seed: every unit a pie of the tasks it wins, thick
lines between the 7 super-clusters. Layers of the middle map (chosen on all children): `r MP[[2]]$layers`.</p>

# 10. Is it new?

Three literature searches (`r 66 + 121 + 80` queries; Semantic Scholar, OpenAlex, Crossref, arXiv, the web; abstracts
and, where open, full texts read; package sources read) for each ingredient:

| Idea | Verdict | Closest published work |
|---|---|---|
| A layered map whose lattice **grows** during training | **partly published** | in kernel form: Wijewardena et al. 2015 grow a GSOM whose similarity is a weighted combination of per-source kernels, a kernel analogue of a growing SuperSOM (abstract only). **Not found** in kohonen's form, with arbitrary per-layer distances (DTW, Manhattan) and pulls in the online scheme. Nearby: the archived R package GrowingSOM (2017) has a growing map with a data and a target layer (fixed weights, sum of squares); multi-stream growing maps keep separate lattices (Nawaratne et al. 2020); `kohonen::expandMap` doubles each side of a *trained* map between runs (Wehrens &amp; Kruisselbrink 2018) |
| Births as the **medoid of the parent's worse-fitting half**, in the multi-layer distance | **not found** | medoid prototypes on fixed maps (Palomino Mariño &amp; de Carvalho 2024); new nodes as the input itself (SOINN, OILFTS 2015) |
| **Robust layers**: Manhattan winner and clipped pull per layer | **not found** as such | robust SOMs (Allende et al. 2004; L&oacute;pez-Rubio et al. 2015) and a per-layer Manhattan distance in kohonen, separately |
| **DTW layers** in a SuperSOM | partly published | the R package `sits` runs an adapted copy of kohonen's supersom with one DTW layer per satellite band (winner by DTW, the usual unwarped Kohonen step); kohonen 3.0's own example puts a shift-tolerant curve distance (WCC) layer next to a sum-of-squares feature layer (Wehrens &amp; Kruisselbrink 2018). New here: DTW next to feature layers, with a warped pull |
| A growing map with **DTW** | published | GSOM Sequence (Fonseka, Alahakoon &amp; Bedingfield, IEEE CIDM 2011) builds DTW into the GSOM; OILFTS (Xu et al. 2015) grows a SOINN network with FastDTW |
| **Learned layer weights** during training | published | Olteanu &amp; Villa-Vialaneix 2015; Palomino Mariño &amp; de Carvalho 2024 (per-view relevance weights) |
| A **label layer** (supervised map) | published | xyf / bdk (Melssen et al. 2006; Wehrens &amp; Buydens 2007); growing semi-supervised GSOMs (Hsu &amp; Halgamuge 2008) |
| A SuperSOM on **wearable IMU** data with feature-set layers | not found | Galea et al. 2021: a supervised super SOM on accelerometer data of cats; human studies use one layer (Mo et al. 2022; Patrikei et al. 2026) |

"Not found" means searched and not found, not proven absent. Two points correct my earlier notes:
growth with DTW is not new (GSOM Sequence, 2011), so my growing DTW-SOM is new only in its medoid births and
their combination; and DTW layers in a SuperSOM exist in software (`sits`). What remains new, as far as we found: the
growing map in kohonen's form with DTW and Manhattan layers and its medoid births; and, as an application, a layered map on
wearable IMU data that recognises children whose gyroscope failed.

# 11. Conclusions

<div class="callout">

* **A layered SOM does not recognise the tasks better than the best single feature set** when every sensor is there:
  `r f3(mn(L_S, "super"))` against `r f3(mn(O_S, "super"))` under the article's rule, a small deficit when each unit names its
  task, both confirmed on fresh seeds. The article's question has a short answer: pick one strong feature set and a robust map.
* **Layering four feature sets with equal weights is worse than the best of them**, because the super-cluster step sees
  the weak layers as much as the strong ones. If layers are used, choose them on the training children.
* **The layered map's real use is a missing sensor**: trained on everything, it recognises children from the
  accelerometer alone `r pt1(ngm("ss_all7") - ngm(best_acc))` points better than a map that only knows the accelerometer,
  and it would have kept the 9 children the article lost.
* **Growing does not change the accuracy here, and DTW lowers it in layered maps**; growing is faster, DTW helps a single
  raw-curve map.

</div>

**Limitations.** One data set (81 children, one study, one sensor position). The article's features were selected
with XGBoost on all 81 children, which flatters every map that uses them on held-out children (not the raw-curve maps).
The raw curves are one choice of representation among many. The missing-sensor gain was measured on 9 children
(72 segments), all from one group of the data set (ids vy01 to vy09), and on a simulation in which `r ac7$better` children
gained and `r ac7$worse` lost.

# 12. Reproduce

```{.bash .fold-show}
Rscript run_supersom.R          # all maps, about 2 hours on 17 cores; results go to supersom/results/
Rscript supersom/tests/test_core.R
```

Then knit `supersom.Rmd` (a few minutes from the cache; `quick: true` knits a small version).

| File | What it holds |
|---|---|
| `supersom/ss_core.cpp` | the layered online SOM: per-layer distances (sum of squares, Manhattan, Euclidean, correlation, DTW) and pulls (clipped, warped), kohonen's winner rule and schedule, the floating radius |
| `supersom/R/ss_maps.R` | layers, kohonen's layer weights, learned weights, the fixed and the growing map, the layer selection |
| `supersom/R/ss_protocols.R` | all data, held-out children, layer vote, mapping with a subset of layers |
| `supersom/R/ss_bench.R` | the 28 maps, the plan, the parallel runs, the cache, the confirmation rule |
| `supersom/R/ss_raw.R` | the posture curves and spectra from the Zenodo file |
| `supersom/R/ss_report.R` | the charts and tables of this page |
| `supersom/tests/test_core.R` | the core against kohonen::som, kohonen::supersom and novel_SOM's DTW |

<details><summary>Sources</summary>

* Wehrens R, Buydens LMC (2007). Self- and super-organizing maps in R: the kohonen package. *J Stat Softw* 21(5).
* Wehrens R, Kruisselbrink J (2018). Flexible self-organizing maps in kohonen 3.0. *J Stat Softw* 87(7).
* Wijewardena M, et al. (2015). Heterogeneous data fusion with multiple kernel growing self organizing maps. *SCAI 2015*, FAIA 278, 167-176.
* Fonseka A, Alahakoon D, Bedingfield S (2011). GSOM sequence: an unsupervised dynamic approach for knowledge discovery in temporal data. *IEEE CIDM*. doi:10.1109/cidm.2011.5949456
* Palomino Mariño LM, de Carvalho FAT (2024). Self-organizing maps with adaptive distances for multiple dissimilarity matrices. *Machine Learning* 113(10), 7783-7806.
* Olteanu M, Villa-Vialaneix N (2015). On-line relational and multiple relational SOM. *Neurocomputing* 147.
* Melssen W, Wehrens R, Buydens L (2006). Supervised Kohonen networks for classification problems. *Chemom Intell Lab Syst* 83.
* Hsu AL, Halgamuge SK (2008). Class structure visualization with semi-supervised growing self-organizing maps. *Neurocomputing* 71.
* Allende H, et al. (2004). Robust self-organizing maps. *CIARP*, LNCS 3287. L&oacute;pez-Rubio E, et al. (2015). Robust self-organization with M-estimators. *Neurocomputing* 151.
* Galea N, Murphy F, Gaschk JL, Schoeman DS, Clemente CJ (2021). Quantifying finer-scale behaviours using self-organising maps (SOMs) to link accelerometery signatures with behavioural patterns in free-roaming terrestrial animals. *Sci Rep* 11, 13566.
* Shen F, Hasegawa O (2006). An incremental network for on-line unsupervised classification and topology learning. *Neural Netw* 19(1).
* Xu H, Xing Y, Shen F, Zhao J (2015). An online incremental learning algorithm for time series. *IJCNN 2015*. doi:10.1109/ijcnn.2015.7280833
* Nawaratne R, Alahakoon D, De Silva D, Kumara H, Yu X (2020). Hierarchical two-stream growing self-organizing maps with transience for human activity recognition. *IEEE Trans Ind Inform* 16(12), 7756-7764.
* Hunziker A (2017). GrowingSOM: Growing Self-Organizing Maps, R package 0.1.1 (CRAN, archived).
* Mo L, Yu H, Hua W (2022). Pruning growing self-organizing map network for human physical activity identification. *J Healthc Eng*.
* `sits`: Satellite image time series analysis, R package (e-sensing, INPE), `R/api_kohonen.R` and `src/kohonen_som.cpp`.
* Patrikei A, Cuberek R, Halfar R, Martinovi&#269; T (2026). Essential time series characteristics for human motion analysis based on Self-Organizing Map clustering. *Acta Gymnica* 56, e2026.004.
* Cuberek R, et al. (2024). Physical exercise measurements with accelerometer and gyroscope. Zenodo, doi:10.5281/zenodo.10984137.

</details>

<details><summary>Session information</summary>

```{r session, echo = FALSE}
sessionInfo()
```

</details>
